### 2013-03-06 Web3D2013-Dec3D/Web3D2013-Dec3DPaper.tex
Jacecs initial version

\documentclass[review]{acmsiggraph}

\TOGonlineid{XXX}
%\title{A Polyfill Approach to Declarative Integration of Interactive 3D Graphics\\ into the World-Wide Web}
\title{Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda}
\author{Name Lastname\thanks{e-mail:name.lastname@someemail.com}\\Some Research Institution}
\pdfauthor{Name Lastname}

\keywords{Declarative 3D, HTML, DOM, Polyfill}

\begin{document}

\maketitle

\begin{abstract}
With the advent of WebGL, plugin-free hardware-accelerated interactive 3D graphics has finally arrived in all major desktop Web browsers.
WebGL is an imperative solution that is tied to the functionality of rasterization API. Consequently, its usage requires a deeper understanding of the rasterization pipeline. In contrast to this stands a declarative approach with an abstract description of the 3D scene. We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related technologies such as XML (that could be used for scene construction), DOM (could be used for scene manipulation), and CSS (could be used for styling), concepts known to the millions of existing Web developers and therefore crucial for the fast adoption of 3D on the Web.

As the participants of the \textit{Declarative 3D for the Web Architecture W3C Community Group}, we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page. In this paper we present Declarative 3D principles that guide the work of the group. We describe the current state of the fundamental to our initiative experimental X3DOM and XML3D frameworks. Finally we draw the agenda for the next development stage of the Declarative 3D for the Web.
\end{abstract}

\begin{CRcatlist}
  \CRcat{I.3.7}{Computer Graphics}{Three-Dimensional Graphics and Realism}{Virtual Reality}
  \CRcat{I.3.6}{Methodology and Techniques}{Standards}{Languages}
\end{CRcatlist}

\keywordlist
\copyrightspace

\section{Introduction}
The Web evolved from a text-based system to the current rich and interactive medium that supports images, 2D graphics, audio and video. These types of new media have made the Web experience richer, more attractive to users, etc, than ever before, and opened up possibilities for new types of applications and usage. The major media type that is still missing is 3D: synthetic, possibly photorealistic images in 3D with animation, as smoothly integrated in the everyday Web experience as images or video. Just as the appearance of images or video could open new application possibilities, access to the 3D on a Web site would make it possible to include realistic models of 3D objects (from models of buildings to representation of the human body or the sceneries for computer games). As of today, such applications require separate applications or the installation of browser plugins; however the goal would be to achieve the same smooth inclusion of 3D content in a Web page like we experience today with images or SVG based 2D graphics.

Although some of these goals could also be achieved by imperative means (e.g., through the usage of WebGL \cite{Khronos-WebGL}), developments of 3D models have a long tradition of using declarative approaches, which is also in line with some of the fundamental principles of Web development. It is therefore important to explore how the experiences accumulated in two different communities, namely the Web Development and Computer Graphics communities, can be capitalized upon to achieve the long term goal of using 3D on the Web the same way as we do with 2D graphics and video today.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{images/Declarative3d.png}
  \caption{The position of the Declarative 3D approach in the current Web graphics technology ecosystem.}
  \label{fig:DeclarativeVsImperative}
  %\vspace{-10px}
\end{figure}

It is arguable that the emerging support for an imperative 3D-API for the Web is useful but insufficient for broad acceptance and usage of 3D on the Web. A declarative approach that is tightly integrated with current web technologies and that offers qualified concepts is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of existing Web developers. The provided concepts must lift the hardware-oriented imperative application programming interfaces (APIs) to an expressive and more easily usable level. Therefore not the low-level data structures of existing hardware layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights. Instead of teaching Web developers 3D graphics APIs, the goal is to bring 3D graphics to the point where its natural for Web developers to just make use of it. While this might not be possible for every possible use of a low-level API, we believe that it can cover the vast majority of use cases.

The \textit{Declarative 3D for the Web Architecture W3C Community Group} \cite{Dec3D-CG} has been formed to suggest and create methods to add interactive high-level declarative 3D objects to the HTML-DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics. The core mission of this Community Group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack which will provide a foundation for future standardization.

This paper presents the current state and efforts of the Dec3D CG. In Section \ref{sec:Principles} we describe Declarative 3D principles: a number of key goals that guide the development of the DOM-based interactive 3D graphics within the Community Group.
In Section \ref{sec:Frameworks} we describe the current state of the fundamental to our initiative experimental systems: Fraunhofer's X3DOM \cite{Behr2009} and DFKI's XML3D \cite{Sons2010} declarative 3D Web publishing frameworks.
In Section \ref{sec:Agenda} we aim to shape the agenda and identify upcoming research issues for the next development stage of the Declarative 3D for the Web. Finally, in Section \ref{sec:Conclusions} we conclude our work.

\section{Declarative 3D Principles}
\label{sec:Principles}
The following goals and principles should guide the development:

\paragraph{Following the Established Principles of the Web}
Declarative 3D is being developed to significantly lower the barrier for authoring 3D content for Web sites. It aims to lower the barrier by duplicating the key features that enabled the growth of the early Web and its further spectacular success.

\begin{description}
  \item [Separation of structure from content] - Underlying the Web from its earliest days was the separation of structure from content. The concept of a paragraph specified by the $<$p$>$ tag was separate from the content in the paragraph. Declarative 3D is attempting to bring the same separation of structure from content to 3D graphics inside of web pages. The concepts such as definitions of 3D objects, transformations, materials, etc. should be implemented in a declarative 3D description as an extension to HTML5 using any existing or future extension mechanism.
  \item [Separation of content from style] - One of the principles of the current Web is also the separation of content and style, most notably through CSS. The successful integration of SVG with HTML was made much easier due to the fact that SVG was already following this principle.
      The aim of the CG is to extend the use of CSS for styling 3D graphics. One example can be the use of the latest CSS3 3D Transforms to allow to manipulate not only 2D but also 3D objects in 3D space.
  \item [Use of the Document Object Model] - The Document Object Model (DOM) is a platform- and language-neutral interface that allows programs and scripts to dynamically access and update the content, structure and style of Web documents \cite{W3C-DOM}. Declarative 3D should use DOM API to examine and modify elements on the 3D scene and their attributes by simply reading and setting their properties.
      As DOM provides access to user actions (e.g., pressing a key or clicking a mouse button), it should also be used as a main interface to interact with 3D content.
\end{description}

Moreover, the CG should reuse existing W3C techniques (specifically from HTML5 and SVG) as far as possible and propose extensions only where 3D specific features are necessary or where they provide significant benefits. Where new concepts are introduced their relation to and effects on existing Web standards should be analyzed, evaluated, and discussed with the respective W3C working groups.

\paragraph{3D Content Creation and Reuse}
While the creation of original 3D geometry and appearances still requires 3D specific know-how, the reuse, configuration, and manipulation of such content should be made similarly easy as for 2D Web content now. The solution should hide internal data structures and algorithms and provide users convenient ways to edit and manipulate such scenes.
A key success factor for Declarative 3D on the Web will be the ability to generate new or reuse existing content. This requires that suitable exporters and converters can be built. However, as 3D on the Web is supposed mainly as a delivery mechanism, it is not necessary to include the ability to semantically represent all but the important 3D features

\paragraph{Platform Independence}
The aim of the CG is to describe 3D content in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only. The technology investigated should allow for content to be portable across user agents, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible. The results of rendering content under such different environments should be highly predictable.

\paragraph{Efficiency and Scalability}
Interactive 3D graphics operates in real-time, which enables new forms of interactivity on the Web but also adds significant new requirements on user agents.  A key requirement for the selected technology therefore is the possibility to implement it efficiently. Since 3D scenes can sometimes become large, any solution should target scalability in the sense that 3D content should run across different platforms (from mobile devices to high-end graphics hardware) with predictable performance. Mechanisms should be in place to handle cases where the performance provided by a user agent on some platform is not sufficient, e.g. by allowing for switching to different content (e.g. lower LOD) or provide alternate methods of delivering the content (e.g. server-based rendering delivered via streaming video).

\paragraph{Security \& Digital Rights Management}
Secure delivery of Web content is a general problem and not specific to 3D data. However, the economic value of 3D data might make the problem more acute. Any proposed solution should therefore be based on a general approach to secure Web content. The CG should, however, collect use cases, extract requirements and examine how far existing methods and standards can be transferred to the proposed architecture. It is already demonstrated that application of XML Encryption and Signature is needed for document fragments as well as full documents, since high-fidelity or sensitive portions of 3D models often need special protections.

\paragraph{Accessibility \& Usability}
A large body of work has shown that accessibility improvements serve all users, not just people with disabilities.  A curious aspect of many 3D graphics approaches is that user navigation is implemented inconsistently.  Therefore, users familiar with one approach are impeded when navigating or interacting with other 3D scenes and models.  Examination of relevant Web Accessibility Initiatives (WAI) principles might provide significant benefit.  Conversely, use of declarative 3D graphics models might provide major benefits when describing the accessibility features and constraints of real-world objects and locations.  Declarative 3D goals and potential solutions may achieve significant benefits if they are harmonizable with WAI imperatives.

\newpage

\section{Declarative 3D Experimental Frameworks}
\label{sec:Frameworks}
As we have already mentioned, our Community Group has been formed to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack which will provide a foundation for future standardization.
While the standardization is our goal, we still need platforms allowing for the experimentation and for the assessment of our design decisions. We also need to reach out to Web developers who could provide us with valuable feedback as early as possible.

In the following we briefly introduce two experimental open-source frameworks that support the ongoing discussion in the computer graphics and Web communities how an integration of HTML and Declarative 3D content could look like. This discussion is followed by the presentation of the \textit{polyfill} approach for the further X3DOM \& XML3D development activities.

\subsection{X3DOM \& XML3D}
Introduction to this section: Fraunhofer's X3DOM \cite{Behr2009} is an open-source framework that allows including X3D \cite{Web3D-X3D} elements as part of any HTML DOM tree... DFKI's XML3D \cite{Sons2010} is a similar approach...

In the second paragraph, I believe we should write something about the main similarities and differences of the frameworks (not going into details but more from philosophical point of view).

In the last paragraph of this section, we can write about the efforts to consolidate approaches and/or how the competing two frameworks influence one another for the common benefit...? This should be followed by the some link to the next section (polyfill).

\newpage

\subsection{Polyfill Approach}
Recently, the CG proposed to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS etc) \cite{Dec3D-LevelsOfIntegration}. The proposal aims to explain existing and possible integration levels and finally to identify an integration level the CG is aiming for.

It quickly became clear that a higher level of integration requires additional APIs in user agents. After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a polyfill approach for the further development activities of the Community Group.

A \textit{polyfill} is downloadable piece of code, which provides facilities that are not built-in to a Web browser \cite{Sharp2010}. For example, many features of HTML5 are not supported by versions of Internet Explorer older than version 8 or 9, but can be used by web pages if those pages install a polyfill. Polyfills can also be used to add entirely new functionality to browsers.

Polyfill based approaches allow the Declarative 3D Community Group to derive hard requirements for related and utilized W3C standards and UA APIs immediately. This leads to a much more evolving concepts and solutions in contrast to an overall declarative 3D specification.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{images/Dec3D-Architecture.png}
  \caption{Declarative 3D polyfill runtime architecture.}
\end{figure}

\begin{figure*}[ht]
  \centering
  \includegraphics[width=\linewidth]{images/Demos.png}
  \caption{ToDo: Some latest examples of X3DOM and XML3D based Web applications.}
  \vspace{-7px}
\end{figure*}

To sum up, the X3DOM and XML3D experimental declarative 3D Web publishing frameworks were designed to explore different options for adding interactive 3D graphics to HTML. We want to stress that the whole integration model is still evolving and it is open for discussion.

\newpage

\section{Declarative 3D Agenda}
\label{sec:Agenda}
During the 1st International Workshop on Declarative 3D for the Web Architecture (Dec3D2012) \cite{Dec3D2012} as well as during more informal group meetings at Web3D in 2011, 2012 and SIGGRAPH in 2011, 2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a declarative 3D standard is:

\paragraph{Encourage Participation}
All relevant stakeholders, e.g. developers, designers, researchers, 3D artists, industry professionals, representatives of standards organizations, accessibility experts, and user-agent implementers, are encouraged to participate in the Declarative 3D Community Group. Participants must be willing to actively develop and donate materials towards the group's deliverables, as well as attend the group's teleconferences and face-to-face meetings.

\paragraph{Clear Definition of Use Cases \& Requirements}
The Community Group needs to agree on a collection of use cases, where embedding 3D data in HTML using declarative approach provides significant benefit. Each use case should explore how publishers and consumers benefit from Dec3D. From these use cases, the Community Group needs to derive and prioritize the different required dimensions for the Declarative 3D technical specification (see \cite{JankowskiDec3D2012,LeFeuvreDec3D2012}).

\paragraph{Clear Technical Specification}
The next step of the CG should be the creation of the clear, detailed and extensible technical specification of the implementation concepts and features necessary to cover the majority of useful requirements (but not necessarily all of them). Measurable properties should be defined to quantitatively or qualitatively evaluate the quality of the solution, document the pros and cons of each solution based on these measurements, demonstrate that, based on the above analysis, there is a good chance of success in creating a W3C standard for Declarative 3D for the Web.

\paragraph{Outreach \& Killer App}
The CG needs to continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the experimental X3DOM and XML3D frameworks. It should also consider focusing on one killer application of Declarative 3D, an application so necessary or desirable that it proves the core value of the Declarative approach and that could substantially increase the visibility of the CG's efforts.

\paragraph{W3C Working Group Proposal}
Finally, the Community Group should deliver a report documenting its progress, any conclusions it arrived at with respect to standardization of \textit{Declarative 3D for the Web} and, if reaching a positive conclusion, recommending a standardization approach as a basis for a future W3C working group on the same topic.

\newpage

\section{Conclusions}
\label{sec:Conclusions}
While WebGL 3D imperative graphics API in the Web context is getting more and more traction, we are still missing an easy way to add interactive high-level declarative 3D objects to the HTML to allow anyone to easily create, share, and experience interactive 3D graphics – with possibly wide ranging effects similar to those caused by the broad availability of video on the Web.

The goal of the Declarative 3D CG is to evaluate the options for a successful standardization of a declarative approach to interactive 3D graphics as part of HTML documents. The goal is to collect suitable use cases, derive requirements from them, and then find the essential set of features and concepts that enables broad uptake by authors and users of interactive 3D on the Web.
We are absolutely aware that our goal is ambitious and it will take some time to implement these features. Therefore, we call for more participation from the 3D Web community that we believe is crucial to achieve our common and ultimate goal: 3D for everyone and everywhere.

%\section*{Acknowledgements}
\bibliographystyle{acmsiggraph}
\bibliography{bibliography}
\end{document}

### 2013-03-06 Web3D2013-Dec3D/paper.tex
new images and updates from Sandy

--- 2013-03-06 Web3D2013-Dec3D/Web3D2013-Dec3DPaper.tex
@@
 \TOGonlineid{XXX}
+
 %\title{A Polyfill Approach to Declarative Integration of Interactive 3D Graphics\\ into the World-Wide Web}
-\title{Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda}
+\title{Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda
+\\~} %some space that we don't get over 4 pages in the final version
 \author{Name Lastname\thanks{e-mail:name.lastname@someemail.com}\\Some Research Institution}
@@
 
-\keywords{Declarative 3D, HTML, DOM, Polyfill}
+\keywords{Declarative 3D, HTML5, DOM Integration, Polyfill}
 
@@
 In contrast to this stands a declarative approach with an abstract description of the 3D scene.
-We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related technologies such as XML (that could be used for scene construction), DOM (could be used for scene manipulation), and CSS (could be used for styling), concepts known to the millions of existing Web developers and therefore crucial for the fast adoption of 3D on the Web.
-
+We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related technologies such as XML (that could be used for scene construction), DOM (could be used for scene manipulation), and CSS (could be used for styling), concepts known to the millions of existing Web developers and therefore crucial for the fast adoption of 3D on the Web.\\
 As the participants of the \textit{Declarative 3D for the Web Architecture W3C Community Group}, we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page.
@@
 
-Although some of these goals could also be achieved by imperative means (e.g., through the usage of WebGL \cite{Khronos-WebGL}), developments of 3D models have a long tradition of using declarative approaches, which is also in line with some of the fundamental principles of Web development.
+Although some of these goals could also be achieved by imperative means (e.g., through the usage of WebGL \cite{WebGL12}), developments of 3D models have a long tradition of using declarative approaches, which is also in line with some of the fundamental principles of Web development.
 It is therefore important to explore how the experiences accumulated in two different communities, namely the Web Development and Computer Graphics communities, can be capitalized upon to achieve the long term goal of using 3D on the Web the same way as we do with 2D graphics and video today.
 
-\begin{figure}[ht]
+\begin{figure}%[ht]
   \centering
@@
 In Section \ref{sec:Principles} we describe Declarative 3D principles: a number of key goals that guide the development of the DOM-based interactive 3D graphics within the Community Group.
-In Section \ref{sec:Frameworks} we describe the current state of the fundamental to our initiative experimental systems: Fraunhofer's X3DOM \cite{Behr2009} and DFKI's XML3D \cite{Sons2010} declarative 3D Web publishing frameworks.
+In Section \ref{sec:Frameworks} we describe the current state of the fundamental to our initiative experimental systems: Fraunhofer IGD's X3DOM \cite{Behr2009} and DFKI's XML3D \cite{Sons2010} declarative 3D Web publishing frameworks.
 In Section \ref{sec:Agenda} we aim to shape the agenda and identify upcoming research issues for the next development stage of the Declarative 3D for the Web.
-Finally, in Section \ref{sec:Conclusions} we conclude our work.
+Finally, in Section \ref{sec:Conclusions} we conclude our work along with directions for future work.
 
@@
 \label{sec:Principles}
+
 The following goals and principles should guide the development:
@@
   \item [Separation of structure from content] - Underlying the Web from its earliest days was the separation of structure from content.
-The concept of a paragraph specified by the $<$p$>$ tag was separate from the content in the paragraph.
+The concept of a paragraph specified by the \verb|<p>| tag was separate from the content in the paragraph.
 Declarative 3D is attempting to bring the same separation of structure from content to 3D graphics inside of web pages.
@@
 
-\newpage
+\paragraph{Leveraging Web Development Infrastructure}
+The Web has become the new application development platform for our time.
+Countless rich internet applications are being created at a dizzying pace.
+The ``apps'' world for mobile applications has also become a huge phenomenon.
+The single most important technology, among many, that has emerged to enable this explosion of creative applications is JavaScript.
+JavaScript has evolved from a toy language into a robust and richly nuanced tool that is the basis for much of the recent explosion in Web applications.
+Declarative 3D, by living within the structure of Web applications is poised to leverage the huge collection of tools and infrastructure that already exists like e.g.
+the JavaScript library jQuery \cite{jquery}.
+The ability to take existing Web debuggers, editors, and viewers and use these tools to create, edit and debug 3D content is a tremendous benefit and just begins to scratch the surface of development tools, created for Web application development, but which we can use for 3D content development.
 
@@
 \label{sec:Frameworks}
+
 As we have already mentioned, our Community Group has been formed to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack which will provide a foundation for future standardization.
@@
 \subsection{X3DOM \& XML3D}
-Introduction to this section: Fraunhofer's X3DOM \cite{Behr2009} is an open-source framework that allows including X3D \cite{Web3D-X3D} elements as part of any HTML DOM tree...
+Introduction to this section: Fraunhofer's X3DOM \cite{Behr2009,Behr12} is an open-source framework that allows including X3D \cite{Web3D-X3D} elements as part of any HTML DOM tree...
 DFKI's XML3D \cite{Sons2010} is a similar approach...
@@
 
-\newpage
+% TODO
+Efficient 3D transmission formats XXX \cite{Behr12}, and are also a major topic at the Khronos Group \cite{Trevett12}.
 
 \subsection{Polyfill Approach}
+
+\begin{figure}%[ht]
+  \centering
+  \includegraphics[width=\linewidth]{images/Dec3D-Architecture.png}
+  \caption{Declarative 3D polyfill runtime architecture.}
+\end{figure}
+
 Recently, the CG proposed to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS etc) \cite{Dec3D-LevelsOfIntegration}.
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\linewidth]{images/Dec3D-Architecture.png}
-  \caption{Declarative 3D polyfill runtime architecture.}
-\end{figure}
-
-\begin{figure*}[ht]
-  \centering
-  \includegraphics[width=\linewidth]{images/Demos.png}
-  \caption{ToDo: Some latest examples of X3DOM and XML3D based Web applications.}
-  \vspace{-7px}
-\end{figure*}
-
 To sum up, the X3DOM and XML3D experimental declarative 3D Web publishing frameworks were designed to explore different options for adding interactive 3D graphics to HTML.
@@
 
-\newpage
+\begin{figure*}%[ht]
+  \centering
+  \includegraphics[width=\linewidth]{images/x3dom_apps.png}
+  \caption{Some X3DOM applications.
+From left to right: walktrough of Roman heritage site; cultural heritage object explorer with metadata; web-based design review application with annotation markers; rendering big data sets, here the Boeing 777, in the Web browser using an ``out-of-browser'' approach.
+(The Boeing 777 commercial airplane model is courtesy of Boeing.
+License granted by David Kasik, Boeing.)}
+  \vspace{-0.2cm}
+\end{figure*}
 
 \section{Declarative 3D Agenda}
+
 \label{sec:Agenda}
@@
 
-\paragraph{Outreach \& Killer App}
+\paragraph{Outreach \& Exemplar Applications}
 The CG needs to continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the experimental X3DOM and XML3D frameworks.
-It should also consider focusing on one killer application of Declarative 3D, an application so necessary or desirable that it proves the core value of the Declarative approach and that could substantially increase the visibility of the CG's efforts.
+%It should also consider focusing on one killer application of Declarative 3D, an application so necessary or desirable that it proves the core value of the Declarative approach and that could substantially increase the visibility of the CG's efforts.
+In this regard, the CG should identify several applications, each requiring and demonstrating different capabilities of Declarative 3D.
+For example, one application could require huge 3D datasets which are impractical for inclusion directly into the DOM; another could require real-time control of an external system (inter process communication); another could integrate with a complex data base; another an illustration of complex data (information visualization) without a physical analog and another a more traditional scientific visualization.
+Key to selection of these applications is the use and demonstration of a Declarative 3D requirement.
 
@@
 
-\newpage
+\section{Conclusions}
 
-\section{Conclusions}
 \label{sec:Conclusions}

### 2013-03-07 Web3D2013-Dec3D/paper.tex
new images

@@
 
-\begin{figure*}%[ht]
+\begin{figure*}
   \centering
-  \includegraphics[width=\linewidth]{images/x3dom_apps.png}
-  \caption{Some X3DOM applications.
-From left to right: walktrough of Roman heritage site; cultural heritage object explorer with metadata; web-based design review application with annotation markers; rendering big data sets, here the Boeing 777, in the Web browser using an ``out-of-browser'' approach.
-(The Boeing 777 commercial airplane model is courtesy of Boeing.
-License granted by David Kasik, Boeing.)}
+  \includegraphics[width=1.0\textwidth]{images/ch_apps.png}
+  \caption{From left to right: walktrough of Roman heritage site; cultural heritage object explorer with metadata; restored virtual synagoge.}
+  \vspace{-0.2cm}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=1.0\textwidth]{images/cad_apps.png}
+  \caption{From left to right: web-based design review application with annotation markers; interactive explosion of CAD model on iPad; lightweight web-based viewer for 3D CAD data; visualization and interactive exploration of simulation data by using extended mouse events.}
   \vspace{-0.2cm}

### 2013-03-07 Web3D2013-Dec3D/paper.tex
some text on x3dom

@@
 \section{Introduction}
+
 The Web evolved from a text-based system to the current rich and interactive medium that supports images, 2D graphics, audio and video.
@@
 It is therefore important to explore how the experiences accumulated in two different communities, namely the Web Development and Computer Graphics communities, can be capitalized upon to achieve the long term goal of using 3D on the Web the same way as we do with 2D graphics and video today.
+Moreover, while imperative graphics APIs are powerful and necessary, a \emph{Declarative 3D} approach can provide web authors an easy way to add interactive high-level declarative 3D objects through the HTML Document Object Model (DOM) allowing them to easily create, modify, share, and experience interactive 3D graphics using HTML documents.
+Figure \ref{fig:DeclarativeVsImperative} depicts the position of the Declarative 3D approach in the current Web graphics technology ecosystem.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{images/Declarative3d.png}
+  \includegraphics[width=1.0\columnwidth]{images/Declarative3d.png}
   \caption{The position of the Declarative 3D approach in the current Web graphics technology ecosystem.}
@@
 
-It is arguable that the emerging support for an imperative 3D-API for the Web is useful but insufficient for broad acceptance and usage of 3D on the Web.
+It is arguable that the emerging support for an imperative 3D API for the Web is useful but insufficient for broad acceptance and usage of 3D on the Web.
 A declarative approach that is tightly integrated with current web technologies and that offers qualified concepts is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of existing Web developers.
@@
 
-The \textit{Declarative 3D for the Web Architecture W3C Community Group} \cite{Dec3D-CG} has been formed to suggest and create methods to add interactive high-level declarative 3D objects to the HTML-DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics.
-The core mission of this Community Group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack which will provide a foundation for future standardization.
+The \textit{Declarative 3D for the Web Architecture W3C Community Group} \cite{Dec3D-CG} was founded to suggest and create methods to add interactive high-level declarative 3D objects to the HTML-DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics.
+The core mission of this Community Group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
+Therefore, this group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality (AR), Efficient XML Interchange (EXI) and other key working groups.
+Certain complex data types and computations are also of mutual interest.
+
+\begin{figure*}%[ht]
+  \centering
+  \includegraphics[width=1.0\textwidth]{images/ch_apps.png}
+  \caption{From left to right: walktrough of Roman heritage site; cultural heritage object explorer with metadata; restored virtual synagoge.}
+  \label{fig:chApps}
+  %\vspace{-0.2cm}
+\end{figure*}
 
@@
 
-\section{Declarative 3D Experimental Frameworks}
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\columnwidth]{images/Dec3D-Architecture.png}
+  \caption{Declarative 3D ``polyfill'' runtime architecture.}
+  \label{fig:polyArch}
+\end{figure}
+
+\section{Declarative 3D Frameworks}
 \label{sec:Frameworks}
@@
 
-In the following we briefly introduce two experimental open-source frameworks that support the ongoing discussion in the computer graphics and Web communities how an integration of HTML and Declarative 3D content could look like.
-This discussion is followed by the presentation of the \textit{polyfill} approach for the further X3DOM \& XML3D development activities.
+In the following we briefly introduce two experimental open-source frameworks that support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
+This discussion is followed by the presentation of the \textit{polyfill} approach for the further X3DOM and XML3D development activities.
 
-\subsection{X3DOM \& XML3D}
-Introduction to this section: Fraunhofer's X3DOM \cite{Behr2009,Behr12} is an open-source framework that allows including X3D \cite{Web3D-X3D} elements as part of any HTML DOM tree...
-DFKI's XML3D \cite{Sons2010} is a similar approach...
+\subsection{X3DOM and XML3D}
 
-In the second paragraph, I believe we should write something about the main similarities and differences of the frameworks (not going into details but more from philosophical point of view).
+Fraunhofer IGD's X3DOM \cite{Behr2009,Behr12} is a JavaScript-based open-source framework for declarative 3D graphics in HTML5 that aims at extending the HTML DOM tree with declarative 3D objects while employing modern Web technologies like CSS3, Ajax, DOM scripting, as well as WebGL and -- as fallback -- Adobe's Flash 11 with Stage 3D for GPU-accelerated rendering.
+The proposed 3D elements are mostly based upon the open ISO standard X3D \cite{Web3D-X3D}, though X3DOM introduces a special HTML profile that basically extends the X3D Interchange profile.
+Additionally, instead of implementing the somehow odd X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.
 
-In the last paragraph of this section, we can write about the efforts to consolidate approaches and/or how the competing two frameworks influence one another for the common benefit...?
+Furthermore, to overcome various problems that come along when embedding 3D mesh data directly into the DOM (which typically consists of several megabytes of vertex data), the developers are working on efficient 3D transmission formats that separate the node structure from the raw vertex data \cite{Behr12}.
+Based on this work, this is now also a major topic at the Khronos Group \cite{Trevett12}, where it is similarly argued that raw 3D data, just like image, video or audio data, needs to be externalized from the HTML document.
+% TODO: finish!!!
+DFKI's XML3D \cite{Sons2010} is a similar approach.
+In contrast to X3DOM, this format proposes a new tag set that is not based on X3D anymore.
+In recent versions of XML3D, mesh data can also be externalized from the HTML document by storing it in separate JSON files.
+
+% TODO
+%XXX In the second paragraph, I believe we should write something about the main similarities and differences of the frameworks (not going into details but more from philosophical point of view).
+% TODO
+%XXX In the last paragraph of this section, we can write about the efforts to consolidate approaches and/or how the competing two frameworks influence one another for the common benefit...?
 This should be followed by the some link to the next section (polyfill).
 
-% TODO
-Efficient 3D transmission formats XXX \cite{Behr12}, and are also a major topic at the Khronos Group \cite{Trevett12}.
+\begin{figure*}
+  \centering
+  \includegraphics[width=1.0\textwidth]{images/cad_apps.png}
+  \caption{From left to right: web-based design review application with annotation markers; interactive explosion of CAD model on iPad; lightweight web-based viewer for 3D CAD data; visualization and interactive exploration of simulation data by using extended mouse events.}
+    \label{fig:cadApps}
+  %\vspace{-0.2cm}
+\end{figure*}
 
 \subsection{Polyfill Approach}
-
-\begin{figure}%[ht]
-  \centering
-  \includegraphics[width=\linewidth]{images/Dec3D-Architecture.png}
-  \caption{Declarative 3D polyfill runtime architecture.}
-\end{figure}
 
@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=1.0\textwidth]{images/ch_apps.png}
-  \caption{From left to right: walktrough of Roman heritage site; cultural heritage object explorer with metadata; restored virtual synagoge.}
-  \vspace{-0.2cm}
-\end{figure*}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=1.0\textwidth]{images/cad_apps.png}
-  \caption{From left to right: web-based design review application with annotation markers; interactive explosion of CAD model on iPad; lightweight web-based viewer for 3D CAD data; visualization and interactive exploration of simulation data by using extended mouse events.}
-  \vspace{-0.2cm}
-\end{figure*}
-
 \section{Declarative 3D Agenda}
@@
 \label{sec:Agenda}
-During the 1st International Workshop on Declarative 3D for the Web Architecture (Dec3D2012) \cite{Dec3D2012} as well as during more informal group meetings at Web3D in 2011, 2012 and SIGGRAPH in 2011, 2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a declarative 3D standard is:
+During the 1st International Workshop on Declarative 3D for the Web Architecture (Dec3D2012) \cite{Dec3D2012} as well as during more informal group meetings at Web3D in 2011, 2012 and SIGGRAPH in 2011, 2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a declarative 3D standard include the following topics.
 

### 2013-03-07 Web3D2013-Dec3D/paper.tex
worked on intro

@@
 
-\TOGonlineid{XXX}
+\TOGonlineid{XXX} %TODO
 
@@
 \\~} %some space that we don't get over 4 pages in the final version
+
 \author{Name Lastname\thanks{e-mail:name.lastname@someemail.com}\\Some Research Institution}
@@
 \begin{abstract}
-With the advent of WebGL, plugin-free hardware-accelerated interactive 3D graphics has finally arrived in all major desktop Web browsers.
-WebGL is an imperative solution that is tied to the functionality of rasterization API.
+With the advent of WebGL, plugin-free hardware-accelerated interactive 3D graphics has finally arrived in all major Web browsers.
+WebGL is an imperative solution that is tied to the functionality of rasterization APIs.
 Consequently, its usage requires a deeper understanding of the rasterization pipeline.
 In contrast to this stands a declarative approach with an abstract description of the 3D scene.
-We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related technologies such as XML (that could be used for scene construction), DOM (could be used for scene manipulation), and CSS (could be used for styling), concepts known to the millions of existing Web developers and therefore crucial for the fast adoption of 3D on the Web.\\
-As the participants of the \textit{Declarative 3D for the Web Architecture W3C Community Group}, we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page.
-In this paper we present Declarative 3D principles that guide the work of the group.
-We describe the current state of the fundamental to our initiative experimental X3DOM and XML3D frameworks.
-Finally we draw the agenda for the next development stage of the Declarative 3D for the Web.
+We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related Web technologies, as those concepts are well-known by millions of Web developers and therefore crucial for the fast adoption of 3D on the Web.%\\
+Hence, in this paper we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page.
+We present declarative 3D principles that guide the work of the \textit{Declarative 3D for the Web Architecture W3C Community Group} and describe the current state of the fundamentals to this initiative.
+Finally, we draw an agenda for the next development stages of Declarative 3D for the Web.
 \end{abstract}
@@
 The major media type that is still missing is 3D: synthetic, possibly photorealistic images in 3D with animation, as smoothly integrated in the everyday Web experience as images or video.
-Just as the appearance of images or video could open new application possibilities, access to the 3D on a Web site would make it possible to include realistic models of 3D objects (from models of buildings to representation of the human body or the sceneries for computer games).
-As of today, such applications require separate applications or the installation of browser plugins; however the goal would be to achieve the same smooth inclusion of 3D content in a Web page like we experience today with images or SVG based 2D graphics.
+Just as the appearance of images or video could open new application possibilities, access to the 3D on a Web site would make it possible to include realistic models of 3D objects -- from models of buildings to representation of the human body or the sceneries for computer games.
+With WebGL \cite{WebGL12}, a JavaScript binding for OpenGL ES 2.0, this seems feasible; however, the goal would be to achieve the same smooth inclusion of 3D content in a Web page like we experience today with images or SVG-based 2D graphics.
 
-Although some of these goals could also be achieved by imperative means (e.g., through the usage of WebGL \cite{WebGL12}), developments of 3D models have a long tradition of using declarative approaches, which is also in line with some of the fundamental principles of Web development.
+Although some of these goals could also be achieved by imperative means (e.g., through the usage of WebGL), developments of 3D models have a long tradition of using declarative approaches, which is also in line with some of the fundamental principles of Web development.
 It is therefore important to explore how the experiences accumulated in two different communities, namely the Web Development and Computer Graphics communities, can be capitalized upon to achieve the long term goal of using 3D on the Web the same way as we do with 2D graphics and video today.
@@
 It is arguable that the emerging support for an imperative 3D API for the Web is useful but insufficient for broad acceptance and usage of 3D on the Web.
-A declarative approach that is tightly integrated with current web technologies and that offers qualified concepts is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of existing Web developers.
+A declarative approach that is tightly integrated with current web technologies, such as JSON or XML (for scene construction), DOM (for scene manipulation), and CSS (for styling), and that offers qualified concepts is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of existing Web developers.
 The provided concepts must lift the hardware-oriented imperative application programming interfaces (APIs) to an expressive and more easily usable level.
 Therefore not the low-level data structures of existing hardware layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights.
-Instead of teaching Web developers 3D graphics APIs, the goal is to bring 3D graphics to the point where its natural for Web developers to just make use of it.
+Instead of teaching Web developers 3D graphics APIs, the goal is to bring 3D graphics to the point where it is natural for Web developers to just make use of it.
 While this might not be possible for every possible use of a low-level API, we believe that it can cover the vast majority of use cases.
 
-The \textit{Declarative 3D for the Web Architecture W3C Community Group} \cite{Dec3D-CG} was founded to suggest and create methods to add interactive high-level declarative 3D objects to the HTML-DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics.
-The core mission of this Community Group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
-Therefore, this group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality (AR), Efficient XML Interchange (EXI) and other key working groups.
-Certain complex data types and computations are also of mutual interest.
+The \textit{Declarative 3D for the Web Architecture W3C Community Group} (Dec3D) \cite{Dec3D-CG} was thus founded to suggest and create methods to add high-level declarative 3D objects to the HTML DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics.
+The core mission of this W3C Community Group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
+Therefore, the group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest.
+In this regard, this paper presents the current state and efforts concerning Dec3D.
 
@@
 
-This paper presents the current state and efforts of the Dec3D CG.
-In Section \ref{sec:Principles} we describe Declarative 3D principles: a number of key goals that guide the development of the DOM-based interactive 3D graphics within the Community Group.
-In Section \ref{sec:Frameworks} we describe the current state of the fundamental to our initiative experimental systems: Fraunhofer IGD's X3DOM \cite{Behr2009} and DFKI's XML3D \cite{Sons2010} declarative 3D Web publishing frameworks.
-In Section \ref{sec:Agenda} we aim to shape the agenda and identify upcoming research issues for the next development stage of the Declarative 3D for the Web.
-Finally, in Section \ref{sec:Conclusions} we conclude our work along with directions for future work.
-
 \section{Declarative 3D Principles}
@@
 
-The following goals and principles should guide the development:
+In this section we describe Declarative 3D principles, where the following key goals should guide the development of DOM-based interactive 3D graphics.
 
@@
 
-In the following we briefly introduce two experimental open-source frameworks that support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
+With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two experimental declarative 3D Web publishing frameworks are available.
+In the following we briefly introduce these open-source frameworks, which support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
 This discussion is followed by the presentation of the \textit{polyfill} approach for the further X3DOM and XML3D development activities.
@@
 
+In this section we aim to shape the agenda and identify upcoming research issues for the next development stage of the Declarative 3D for the Web.
+
 \label{sec:Agenda}

### 2013-03-08 Web3D2013-Dec3D/paper.tex
improvements

@@
   \label{fig:DeclarativeVsImperative}
-  %\vspace{-10px}
+  \vspace{-0.2cm}
 \end{figure}
@@
 
-The \textit{Declarative 3D for the Web Architecture W3C Community Group} (Dec3D) \cite{Dec3D-CG} was thus founded to suggest and create methods to add high-level declarative 3D objects to the HTML DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics.
-The core mission of this W3C Community Group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
+The \textit{Declarative 3D for the Web Architecture W3C Community Group} (Dec3D) \cite{Dec3D} was thus founded to suggest and create methods to add high-level declarative 3D objects to the HTML DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics.
+The core mission of this group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
 Therefore, the group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest.
@@
   \label{fig:chApps}
-  %\vspace{-0.2cm}
+  \vspace{-0.2cm}
 \end{figure*}
@@
 
-In this section we describe Declarative 3D principles, where the following key goals should guide the development of DOM-based interactive 3D graphics.
+Here we describe declarative 3D principles, where the following goals should guide the development of DOM-based 3D graphics.
 
 \paragraph{Following the Established Principles of the Web}
-Declarative 3D is being developed to significantly lower the barrier for authoring 3D content for Web sites.
-It aims to lower the barrier by duplicating the key features that enabled the growth of the early Web and its further spectacular success.
+Declarative 3D is being developed to significantly lower the barrier for authoring 3D content for Web sites by duplicating the key features that enabled the growth of the early Web and its further success.
 
 \begin{description}
-  \item [Separation of structure from content] - Underlying the Web from its earliest days was the separation of structure from content.
-The concept of a paragraph specified by the \verb|<p>| tag was separate from the content in the paragraph.
-Declarative 3D is attempting to bring the same separation of structure from content to 3D graphics inside of web pages.
+  \item [Separation of structure from content] Underlying the Web from its earliest days was the separation of structure from content.
+The concept of a paragraph specified by the \verb|<p>| tag was separate from its content.
+Declarative 3D is attempting to bring the same separation to 3D graphics inside of web pages.
 The concepts such as definitions of 3D objects, transformations, materials, etc.
-should be implemented in a declarative 3D description as an extension to HTML5 using any existing or future extension mechanism.
-  \item [Separation of content from style] - One of the principles of the current Web is also the separation of content and style, most notably through CSS.
-The successful integration of SVG with HTML was made much easier due to the fact that SVG was already following this principle.
-      The aim of the CG is to extend the use of CSS for styling 3D graphics.
-One example can be the use of the latest CSS3 3D Transforms to allow to manipulate not only 2D but also 3D objects in 3D space.
-  \item [Use of the Document Object Model] - The Document Object Model (DOM) is a platform- and language-neutral interface that allows programs and scripts to dynamically access and update the content, structure and style of Web documents \cite{W3C-DOM}.
-Declarative 3D should use DOM API to examine and modify elements on the 3D scene and their attributes by simply reading and setting their properties.
-      As DOM provides access to user actions (e.g., pressing a key or clicking a mouse button), it should also be used as a main interface to interact with 3D content.
+should be implemented in a declarative 3D description as an extension to HTML5 \cite{HTML5} using any existing or future extension mechanism.
+  \item [Separation of content from style] One of the principles of the current Web is also the separation of content and style, most notably through CSS.
+The successful integration of SVG \cite{svg} with HTML was made much easier due to the fact that SVG was already following this principle.
+The objective here is to extend the use of CSS for styling 3D graphics.
+One example can be the use of the latest CSS3 3D Transforms \cite{webkit3DCSS} to allow manipulating not only 2D but also 3D objects.
+  \item [Use of the Document Object Model] The DOM \cite{W3C-DOM} is a platform- and language-neutral interface that allows programs and scripts to dynamically access and update the content, structure and style of Web documents.
+Declarative 3D should use the DOM API to examine and modify elements on the 3D scene and their attributes by simply reading and setting their properties.
+As the DOM provides access to user actions (e.g., pressing a key or clicking a mouse button), it should also be used as a main interface to interact with 3D contents.
 \end{description}
 
-Moreover, the CG should reuse existing W3C techniques (specifically from HTML5 and SVG) as far as possible and propose extensions only where 3D specific features are necessary or where they provide significant benefits.
+Moreover, embedded 3D graphics should reuse existing W3C techniques (specifically from HTML5 and SVG) as far as possible and propose extensions only where 3D-specific features are necessary or where they provide significant benefits.
 Where new concepts are introduced their relation to and effects on existing Web standards should be analyzed, evaluated, and discussed with the respective W3C working groups.
@@
 This requires that suitable exporters and converters can be built.
-However, as 3D on the Web is supposed mainly as a delivery mechanism, it is not necessary to include the ability to semantically represent all but the important 3D features
+However, as 3D on the Web is supposed mainly as a delivery mechanism, it is not necessary to include the ability to semantically represent all 3D features.
 
 \paragraph{Platform Independence}
-The aim of the CG is to describe 3D content in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only.
+Another aspect is to describe 3D content in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only (cp.
+\cite{Schwenk12}).
 The technology investigated should allow for content to be portable across user agents, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible.
@@
 \paragraph{Efficiency and Scalability}
-Interactive 3D graphics operates in real-time, which enables new forms of interactivity on the Web but also adds significant new requirements on user agents.
-A key requirement for the selected technology therefore is the possibility to implement it efficiently.
-Since 3D scenes can sometimes become large, any solution should target scalability in the sense that 3D content should run across different platforms (from mobile devices to high-end graphics hardware) with predictable performance.
+Interactive real-time 3D graphics enables new forms of interactivity on the Web but also adds significant new requirements on user agents.
+A key requirement for the selected technology therefore is the possibility to implement it efficiently (cf.
+e.g.
+\cite{Trevett12}).
+Since 3D scenes can become rather large, any solution should target scalability in the sense that 3D content should run across different platforms (from mobile devices to high-end graphics hardware) with predictable performance.
 Mechanisms should be in place to handle cases where the performance provided by a user agent on some platform is not sufficient, e.g.
@@
 
-\paragraph{Security \& Digital Rights Management}
+\paragraph{Security and Digital Rights Management}
 Secure delivery of Web content is a general problem and not specific to 3D data.
@@
 Any proposed solution should therefore be based on a general approach to secure Web content.
-The CG should, however, collect use cases, extract requirements and examine how far existing methods and standards can be transferred to the proposed architecture.
-It is already demonstrated that application of XML Encryption and Signature is needed for document fragments as well as full documents, since high-fidelity or sensitive portions of 3D models often need special protections.
+However, we first need to collect use cases, extract requirements and examine how far existing methods (e.g.
+\cite{KollerL05}) and standards can be transferred to the proposed architecture.
+It is already demonstrated that the application of XML Encryption and Signature is needed for document fragments as well as full documents, since high-fidelity or sensitive portions of 3D models often need special protections.
 
-\paragraph{Accessibility \& Usability}
+\paragraph{Accessibility and Usability}
 A large body of work has shown that accessibility improvements serve all users, not just people with disabilities.
-A curious aspect of many 3D graphics approaches is that user navigation is implemented inconsistently.
+A problematic aspect of many 3D graphics approaches however is that user navigation and interaction is implemented inconsistently.
 Therefore, users familiar with one approach are impeded when navigating or interacting with other 3D scenes and models.
@@
 The ``apps'' world for mobile applications has also become a huge phenomenon.
-The single most important technology, among many, that has emerged to enable this explosion of creative applications is JavaScript.
-JavaScript has evolved from a toy language into a robust and richly nuanced tool that is the basis for much of the recent explosion in Web applications.
-Declarative 3D, by living within the structure of Web applications is poised to leverage the huge collection of tools and infrastructure that already exists like e.g.
-the JavaScript library jQuery \cite{jquery}.
-The ability to take existing Web debuggers, editors, and viewers and use these tools to create, edit and debug 3D content is a tremendous benefit and just begins to scratch the surface of development tools, created for Web application development, but which we can use for 3D content development.
+One of the most important enabling technologies is JavaScript \cite{Crockford08}, which has evolved from a toy language into a robust and richly nuanced tool that is the basis for much of the recent explosion in Web applications.
+Declarative 3D, by living within the structure of Web applications is poised to leverage the huge collection of tools and infrastructure that already exists such as the JavaScript library jQuery \cite{jquery}.
+The ability to take existing Web debuggers, editors, and viewers and use these tools to create, edit, and debug 3D content is a tremendous benefit and just begins to scratch the surface of development tools, created for Web application development, but which we can use for 3D content development.
 
@@
   \includegraphics[width=1.0\columnwidth]{images/Dec3D-Architecture.png}
-  \caption{Declarative 3D ``polyfill'' runtime architecture.}
+  \caption{Proposed declarative 3D ``polyfill'' runtime architecture.}
   \label{fig:polyArch}
+  \vspace{-0.2cm}
 \end{figure}
@@
 
-As we have already mentioned, our Community Group has been formed to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack which will provide a foundation for future standardization.
-While the standardization is our goal, we still need platforms allowing for the experimentation and for the assessment of our design decisions.
+As already mentioned, the Dec3D Community Group \cite{Dec3D} has been formed to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack, which will provide a foundation for future standardization.
+While this standardization is our goal, we still need platforms allowing for the experimentation and for the assessment of our design decisions.
 We also need to reach out to Web developers who could provide us with valuable feedback as early as possible.
@@
 In the following we briefly introduce these open-source frameworks, which support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
-This discussion is followed by the presentation of the \textit{polyfill} approach for the further X3DOM and XML3D development activities.
+This discussion is followed by the presentation of the so-called ``polyfill'' approach for the further X3DOM and XML3D development activities.
 
@@
 
-Fraunhofer IGD's X3DOM \cite{Behr2009,Behr12} is a JavaScript-based open-source framework for declarative 3D graphics in HTML5 that aims at extending the HTML DOM tree with declarative 3D objects while employing modern Web technologies like CSS3, Ajax, DOM scripting, as well as WebGL and -- as fallback -- Adobe's Flash 11 with Stage 3D for GPU-accelerated rendering.
+Fraunhofer IGD's X3DOM \cite{Behr2009,Behr12} is a JavaScript-based open-source framework for declarative 3D graphics in HTML5 that aims at extending the HTML DOM tree with declarative 3D objects while employing modern Web technologies like CSS3, Ajax, DOM scripting, as well as WebGL and -- as fallback -- Adobe's Flash 11 with Stage 3D \cite{flash} for GPU-accelerated rendering.
 The proposed 3D elements are mostly based upon the open ISO standard X3D \cite{Web3D-X3D}, though X3DOM introduces a special HTML profile that basically extends the X3D Interchange profile.
@@
   \label{fig:cadApps}
-  %\vspace{-0.2cm}
+  \vspace{-0.2cm}
 \end{figure*}
@@
 
-In this section we aim to shape the agenda and identify upcoming research issues for the next development stage of the Declarative 3D for the Web.
+In this section we shape the agenda and identify upcoming research issues for the next development stage of Dec3D.
 
 \label{sec:Agenda}
-During the 1st International Workshop on Declarative 3D for the Web Architecture (Dec3D2012) \cite{Dec3D2012} as well as during more informal group meetings at Web3D in 2011, 2012 and SIGGRAPH in 2011, 2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a declarative 3D standard include the following topics.
+During the 1st International Workshop on Declarative 3D for the Web Architecture (Dec3D2012) \cite{Dec3D2012} as well as during more informal group meetings at Web3D and SIGGRAPH in 2011 and 2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a declarative 3D standard include the following topics.
 
@@
 
-\paragraph{Clear Definition of Use Cases \& Requirements}
+\paragraph{Clear Definition of Use Cases and Requirements}
 The Community Group needs to agree on a collection of use cases, where embedding 3D data in HTML using declarative approach provides significant benefit.
@@
 
-\paragraph{Outreach \& Exemplar Applications}
+\paragraph{Outreach and Exemplar Applications}
 The CG needs to continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the experimental X3DOM and XML3D frameworks.
@@
 \section{Conclusions}
+\label{sec:Conclusions}
 
-\label{sec:Conclusions}
-While WebGL 3D imperative graphics API in the Web context is getting more and more traction, we are still missing an easy way to add interactive high-level declarative 3D objects to the HTML to allow anyone to easily create, share, and experience interactive 3D graphics – with possibly wide ranging effects similar to those caused by the broad availability of video on the Web.
+While WebGL, a 3D imperative graphics API in the Web context, is getting more and more traction, we are still missing an easy way to add interactive high-level declarative 3D objects to an HTML document to allow anyone to easily create, share, and experience interactive 3D graphics, with possibly wide ranging effects similar to those caused by the broad availability of video on the Web.
 
-The goal of the Declarative 3D CG is to evaluate the options for a successful standardization of a declarative approach to interactive 3D graphics as part of HTML documents.
-The goal is to collect suitable use cases, derive requirements from them, and then find the essential set of features and concepts that enables broad uptake by authors and users of interactive 3D on the Web.
+The objective of this position paper on Declarative 3D is to evaluate the options for a successful standardization of a declarative approach to interactive 3D graphics as part of HTML documents.
+The idea is to collect suitable use cases, derive requirements from them, and then find the essential set of features and concepts that enables broad uptake by authors and users of interactive 3D on the Web.
 We are absolutely aware that our goal is ambitious and it will take some time to implement these features.
-Therefore, we call for more participation from the 3D Web community that we believe is crucial to achieve our common and ultimate goal: 3D for everyone and everywhere.
+Therefore, we call for more participation from the Web3D and W3C communities that we believe is crucial to achieve our common and ultimate goal: 3D for everyone and everywhere.
 

### 2013-03-08 Web3D2013-Dec3D/paper.tex
more text

@@
 \title{Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda
-\\~} %some space that we don't get over 4 pages in the final version
+\\~} %some space that we don't get over 4(?) pages in the final version
 
@@
 In contrast to this stands a declarative approach with an abstract description of the 3D scene.
-We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related Web technologies, as those concepts are well-known by millions of Web developers and therefore crucial for the fast adoption of 3D on the Web.%\\
+We strongly believe that such approach is more suitable for the integration of 3D into HTML5 and related Web technologies, as those concepts are well-known by millions of Web developers and therefore crucial for the fast adoption of 3D on the Web.
 Hence, in this paper we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page.
@@
 It is arguable that the emerging support for an imperative 3D API for the Web is useful but insufficient for broad acceptance and usage of 3D on the Web.
-A declarative approach that is tightly integrated with current web technologies, such as JSON or XML (for scene construction), DOM (for scene manipulation), and CSS (for styling), and that offers qualified concepts is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of existing Web developers.
-The provided concepts must lift the hardware-oriented imperative application programming interfaces (APIs) to an expressive and more easily usable level.
-Therefore not the low-level data structures of existing hardware layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights.
+A declarative approach that offers qualified concepts and that is tightly integrated with current web technologies, such as JSON or XML for scene construction, DOM for scene manipulation, and CSS for styling, is necessary to support a fast adoption and broad use of interactive 3D graphics by the millions of Web developers.
+The provided concepts must lift the hardware-oriented imperative application programming interfaces to an expressive and more easily usable level.
+Therefore not the low-level data structures of existing GPU layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights.
 Instead of teaching Web developers 3D graphics APIs, the goal is to bring 3D graphics to the point where it is natural for Web developers to just make use of it.
-While this might not be possible for every possible use of a low-level API, we believe that it can cover the vast majority of use cases.
+While this might not be possible for every possible use of a low-level API, it still can cover the vast majority of use cases.
 
-The \textit{Declarative 3D for the Web Architecture W3C Community Group} (Dec3D) \cite{Dec3D} was thus founded to suggest and create methods to add high-level declarative 3D objects to the HTML DOM \cite{W3C-DOM}, so users can easily create, share and experience interactive 3D graphics.
-The core mission of this group is to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
-Therefore, the group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest.
+The \textit{Declarative 3D for the Web Architecture W3C Community Group} (Dec3D) \cite{Dec3D} was thus founded to suggest and create methods to add high-level declarative 3D objects to the HTML DOM \cite{W3C-DOM}, so users can share and experience interactive 3D contents.
+Moreover, this not only allows creating new content from existing content but also to index and search 3D content.
+The core mission is to determine the requirements, options, and use cases for the integration of interactive 3D graphics into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
+Therefore, the group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) \cite{EXI} and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest.
 In this regard, this paper presents the current state and efforts concerning Dec3D.
@@
 \cite{Trevett12}).
+Native implementations allow utilizing all (battery) resources more efficiently while leveraging heterogeneous hardware.
+Thereby, the CPU time can be used for the application instead for rendering, collision, scene-housekeeping, etc., which is esp.
+critical for mobile devices.
 Since 3D scenes can become rather large, any solution should target scalability in the sense that 3D content should run across different platforms (from mobile devices to high-end graphics hardware) with predictable performance.
@@
 \paragraph{Accessibility and Usability}
-A large body of work has shown that accessibility improvements serve all users, not just people with disabilities.
+Accessibility improvements serve all users, not just people with disabilities.
 A problematic aspect of many 3D graphics approaches however is that user navigation and interaction is implemented inconsistently.
@@
 The ``apps'' world for mobile applications has also become a huge phenomenon.
-One of the most important enabling technologies is JavaScript \cite{Crockford08}, which has evolved from a toy language into a robust and richly nuanced tool that is the basis for much of the recent explosion in Web applications.
-Declarative 3D, by living within the structure of Web applications is poised to leverage the huge collection of tools and infrastructure that already exists such as the JavaScript library jQuery \cite{jquery}.
+One important enabling technologies is JavaScript \cite{Crockford08}, which has evolved from a toy language into a robust and richly nuanced tool that is the basis for much of the recent explosion in Web applications.
+Declarative 3D, by living within the structure of Web apps is poised to leverage the huge collection of tools and infrastructure that already exists such as jQuery \cite{jquery} etc.
 The ability to take existing Web debuggers, editors, and viewers and use these tools to create, edit, and debug 3D content is a tremendous benefit and just begins to scratch the surface of development tools, created for Web application development, but which we can use for 3D content development.
@@
 
+\subsection{X3DOM and XML3D}
+
 With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two experimental declarative 3D Web publishing frameworks are available.
@@
 This discussion is followed by the presentation of the so-called ``polyfill'' approach for the further X3DOM and XML3D development activities.
-
-\subsection{X3DOM and XML3D}
 
@@
 
-Furthermore, to overcome various problems that come along when embedding 3D mesh data directly into the DOM (which typically consists of several megabytes of vertex data), the developers are working on efficient 3D transmission formats that separate the node structure from the raw vertex data \cite{Behr12}.
-Based on this work, this is now also a major topic at the Khronos Group \cite{Trevett12}, where it is similarly argued that raw 3D data, just like image, video or audio data, needs to be externalized from the HTML document.
-% TODO: finish!!!
+Furthermore, to overcome various problems that come along when embedding 3D mesh data, which typically consists of several megabytes of vertex attribute data, directly into the DOM tree, the developers are working on efficient 3D mesh encodings that allow separating the node structure from the raw vertex data \cite{Behr12}.
+In this regard, Figures \ref{fig:chApps} and \ref{fig:cadApps} show several X3DOM applications from the cultural heritage and engineering domain, where the ability to efficiently handle big 3D data sets is of high importance.
+Based on this work, 3D transmission formats are now also a major topic at the Khronos Group \cite{Trevett12}, where it is similarly argued that raw 3D data, just like image, video or audio data, needs to be externalized from the HTML document.
+
 DFKI's XML3D \cite{Sons2010} is a similar approach.
 In contrast to X3DOM, this format proposes a new tag set that is not based on X3D anymore.
+Furthermore, it does not define a scene-graph but a strict tree structure by means of a prototyping mechanism.
 In recent versions of XML3D, mesh data can also be externalized from the HTML document by storing it in separate JSON files.
+Moreover, the developers are working on a system for dynamic data-flow to efficiently support animations \cite{KleinSJRSB12}.
 
-% TODO
-%XXX In the second paragraph, I believe we should write something about the main similarities and differences of the frameworks (not going into details but more from philosophical point of view).
-% TODO
-%XXX In the last paragraph of this section, we can write about the efforts to consolidate approaches and/or how the competing two frameworks influence one another for the common benefit...?
+% TODO: In the last paragraph of this section, we can write about the efforts to consolidate approaches and/or how the competing two frameworks influence one another for the common benefit...?
 This should be followed by the some link to the next section (polyfill).
@@
 
-Recently, the CG proposed to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS etc) \cite{Dec3D-LevelsOfIntegration}.
-The proposal aims to explain existing and possible integration levels and finally to identify an integration level the CG is aiming for.
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\columnwidth]{images/polyfill.png}
+  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
+  \label{fig:polyInt}
+  \vspace{-0.2cm}
+\end{figure}
 
+% TODO: don't cite this link but present the whole idea here!!!
+% http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
+Recently, the group proposed to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS.
+etc.).
+The proposal aims to explain existing and possible integration levels (Figure \ref{fig:polyInt}) and finally to identify an integration level the group is aiming for.
+% again, this is still bla bla
+% instead, insert text from slides - see slide 22 ff.
 It quickly became clear that a higher level of integration requires additional APIs in user agents.
-After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a polyfill approach for the further development activities of the Community Group.
+After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a polyfill approach for the further development activities.
+Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.
 
-A \textit{polyfill} is downloadable piece of code, which provides facilities that are not built-in to a Web browser \cite{Sharp2010}.
-For example, many features of HTML5 are not supported by versions of Internet Explorer older than version 8 or 9, but can be used by web pages if those pages install a polyfill.
+A \textit{polyfill} is a downloadable piece of code, which provides facilities that are not natively built-in to a Web browser \cite{Sharp2010}.
+For example, many features of HTML5 are not supported by versions of Internet Explorer older than version 8 or 9, but can be used by web pages if those pages install a so-called polyfill.
 Polyfills can also be used to add entirely new functionality to browsers.
 
-Polyfill based approaches allow the Declarative 3D Community Group to derive hard requirements for related and utilized W3C standards and UA APIs immediately.
-This leads to a much more evolving concepts and solutions in contrast to an overall declarative 3D specification.
+Polyfill-based approaches allow deriving hard requirements for related and utilized W3C standards and user agent APIs immediately.
+This leads to much more evolving concepts and solutions in contrast to an overall declarative 3D specification.
+To sum up, the X3DOM and XML3D experimental declarative 3D Web publishing frameworks are designed to explore different options for adding 3D graphics to HTML.
+Here, Figure \ref{fig:polyArch} depicts our proposed declarative 3D polyfill runtime architecture, though we'd like to stress that the whole integration model is still evolving and open for discussion.
 
-To sum up, the X3DOM and XML3D experimental declarative 3D Web publishing frameworks were designed to explore different options for adding interactive 3D graphics to HTML.
-We want to stress that the whole integration model is still evolving and it is open for discussion.
+\subsection{Declarative 3D Essentials}
+
+TODO % see slides, where these are discussed starting with slide 11/12
+
+Dec3D Scene Graph
+Modern and minimalistic scene graph or tree
+traversal and culling
+creates mesh/material/transform bundles for render layer
+picking
+Supports graph and tree structures for data reuse
+API layer, no DOM integration!
+
+Essential Dec3D DOM Elements
+Focus on minimalistic element set, not usability
+Represents SG concepts as DOM elements
+Adds DOM, Events and CSS to SG
+
+Comprehensive Dec3D DOM Elements
+Builds on Essential and/or SG layer
+Focus on usability not minimalism
+Could build on existing SG standards (e.g. X3D) for convenience
+
+1. Start element
+Transition from HTML flow layout to 3D transform
+
+2. CSS integration
+Use exiting CSS properties where applicable
+Define a set of 3D-specific CSS properties
+Needs to be supported in UAs!
+
+3. Hierarchy
+HTML hierarchy
+Comes with issues (e.g. CSS inheritance)
+
+4. Transformations
+Use CSS 3D Transforms
+
+5. Generic Data Containers
+Take modern HW into account
+Name: Type, TupleSize, Value
+Based on TypedArrays with interface to modify data
+
+6. External Data Containers
+External (binary) blobs / transmission format
+Maps to generic data containers
+
+7. Operators on Data Containers
+Allows dynamic meshes, image processing, compression
+Data driven approach: Xflow, X3D Routes, SVG Filter FX
+
+8. Shaders
+2-3 powerful fixed-function shaders
+Enough for many use-cases
+No timing attack issues
+Can adapt to target device
+
+9. Textures
+Use HTML elements: \verb|<img>|, \verb|<video>|, \verb|<canvas>|
+Interactive textures: \verb|<svg>|, \verb|<html>|
+
+10. 3D Geometry?
+Close to OpenGL primitives
+
+11. Lights
+Shader parameters in GL, but don't obstruct future concepts
+Intuitive way to define lights
+
+12. Viewpoints?
+Link to viewpoints (intra and inter-document)
+
+13. Camera navigation
+Pre-defined navigation modes?
+
+14. DOM Events
+Support standard events: mouseover, click
+3D-specific DOM Events contain 3D-specific context information
+
+15. Interfaces and Types?
+Provide IDLs for new elements
+Convenience functions
+Dec3D Base Types: Lean set of complex data types for 2D/3D
+Provide / use types that help to manage 3D scene, e.g. CSS Matrix
 
 \section{Declarative 3D Agenda}
+\label{sec:Agenda}
 
 In this section we shape the agenda and identify upcoming research issues for the next development stage of Dec3D.
-
-\label{sec:Agenda}
-During the 1st International Workshop on Declarative 3D for the Web Architecture (Dec3D2012) \cite{Dec3D2012} as well as during more informal group meetings at Web3D and SIGGRAPH in 2011 and 2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a declarative 3D standard include the following topics.
+During the 1\textsuperscript{st} International Workshop on Declarative 3D for the Web Architecture \cite{Dec3D2012} as well as during more informal group meetings at Web3D and SIGGRAPH in 2011/2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a Dec3D standard include the following topics.
 
@@
 All relevant stakeholders, e.g.
-developers, designers, researchers, 3D artists, industry professionals, representatives of standards organizations, accessibility experts, and user-agent implementers, are encouraged to participate in the Declarative 3D Community Group.
-Participants must be willing to actively develop and donate materials towards the group's deliverables, as well as attend the group's teleconferences and face-to-face meetings.
+developers, designers, researchers, 3D artists, industry professionals, representatives of standards organizations, accessibility experts, and user-agent implementers, are encouraged to participate in this group.
+Participants should be willing to actively develop and donate materials towards the group's deliverables, as well as attend the group's teleconferences and face-to-face meetings.
 
 \paragraph{Clear Definition of Use Cases and Requirements}
-The Community Group needs to agree on a collection of use cases, where embedding 3D data in HTML using declarative approach provides significant benefit.
+The group needs to agree on a collection of use cases, where embedding 3D data in HTML using a declarative approach provides significant benefit.
+Here, declarative approaches esp.
+seem relevant for the industry, since they tend to think in formats not in APIs.
 Each use case should explore how publishers and consumers benefit from Dec3D.
-From these use cases, the Community Group needs to derive and prioritize the different required dimensions for the Declarative 3D technical specification (see \cite{JankowskiDec3D2012,LeFeuvreDec3D2012}).
+From these use cases, the group needs to derive and prioritize the different required dimensions for the Dec3D technical specification \cite{JankowskiDec3D2012,LeFeuvreDec3D2012}.
 
 \paragraph{Clear Technical Specification}
-The next step of the CG should be the creation of the clear, detailed and extensible technical specification of the implementation concepts and features necessary to cover the majority of useful requirements (but not necessarily all of them).
-Measurable properties should be defined to quantitatively or qualitatively evaluate the quality of the solution, document the pros and cons of each solution based on these measurements, demonstrate that, based on the above analysis, there is a good chance of success in creating a W3C standard for Declarative 3D for the Web.
+The next step is the creation of the clear, detailed and extensible technical specification of the implementation concepts and features necessary to cover a majority of useful requirements.
+Measurable properties need to be defined to quantitatively and/or qualitatively evaluate the achieved solution, document the pros and cons of each, and demonstrate that, based on the above analysis, there is a good chance of success in creating a W3C standard for Declarative 3D for the Web.
 
 \paragraph{Outreach and Exemplar Applications}
-The CG needs to continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the experimental X3DOM and XML3D frameworks.
-%It should also consider focusing on one killer application of Declarative 3D, an application so necessary or desirable that it proves the core value of the Declarative approach and that could substantially increase the visibility of the CG's efforts.
-In this regard, the CG should identify several applications, each requiring and demonstrating different capabilities of Declarative 3D.
-For example, one application could require huge 3D datasets which are impractical for inclusion directly into the DOM; another could require real-time control of an external system (inter process communication); another could integrate with a complex data base; another an illustration of complex data (information visualization) without a physical analog and another a more traditional scientific visualization.
-Key to selection of these applications is the use and demonstration of a Declarative 3D requirement.
+The group needs to continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the open-source frameworks X3DOM and XML3D.
+Therefore, several applications should be identified, each requiring and demonstrating different capabilities of Dec3D.
+For example, one application could require huge 3D datasets which are impractical for inclusion directly into the DOM; another could require real-time control of an external system (inter-process communication); another could integrate with a complex data base; another an illustration of complex data (information visualization) without a physical analog, and another a more traditional scientific visualization.
+Key to selection of these applications is the use and demonstration of a declarative 3D requirement.
 
 \paragraph{W3C Working Group Proposal}
-Finally, the Community Group should deliver a report documenting its progress, any conclusions it arrived at with respect to standardization of \textit{Declarative 3D for the Web} and, if reaching a positive conclusion, recommending a standardization approach as a basis for a future W3C working group on the same topic.
+Finally, the Community Group should deliver reports documenting its progress, any conclusions it arrived at with respect to the standardization of \textit{Declarative 3D for the Web} and, if reaching a positive conclusion, recommending a standardization approach as a basis for a future W3C working group on the same topic.
 
@@
 While WebGL, a 3D imperative graphics API in the Web context, is getting more and more traction, we are still missing an easy way to add interactive high-level declarative 3D objects to an HTML document to allow anyone to easily create, share, and experience interactive 3D graphics, with possibly wide ranging effects similar to those caused by the broad availability of video on the Web.
+The main motivation is thus to make it easy to add 3D to Web pages by bringing 3D to the Web developers and not vice versa.
+This can be achieved by fully integrating 3D content into HTML5 documents, where interactive 3D graphics is a first class DOM objects.
+Moreover, by reusing existing Web technology wherever possible, no new concepts are added unless absolutely necessary.
 
-The objective of this position paper on Declarative 3D is to evaluate the options for a successful standardization of a declarative approach to interactive 3D graphics as part of HTML documents.
+Another objective of this position paper on Declarative 3D for the Web is to evaluate the options for a successful standardization of a declarative approach to interactive 3D graphics as part of HTML documents.
 The idea is to collect suitable use cases, derive requirements from them, and then find the essential set of features and concepts that enables broad uptake by authors and users of interactive 3D on the Web.
@@
 %\section*{Acknowledgements}
+
 \bibliographystyle{acmsiggraph}
+%\parskip3pt
 \bibliography{bibliography}
+
 \end{document}

### 2013-03-09 Web3D2013-Dec3D/paper.tex
finished 1st version

@@
 
-Moreover, embedded 3D graphics should reuse existing W3C techniques (specifically from HTML5 and SVG) as far as possible and propose extensions only where 3D-specific features are necessary or where they provide significant benefits.
+Embedded 3D graphics should reuse existing W3C techniques, specifically from HTML5 and SVG, as far as possible and propose extensions only where specific features are necessary or provide significant benefits.
 Where new concepts are introduced their relation to and effects on existing Web standards should be analyzed, evaluated, and discussed with the respective W3C working groups.
@@
 \paragraph{Platform Independence}
-Another aspect is to describe 3D content in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only (cp.
-\cite{Schwenk12}).
-The technology investigated should allow for content to be portable across user agents, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible.
+3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only \cite{Schwenk12}.
+This should allow for content to be portable across browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible.
 The results of rendering content under such different environments should be highly predictable.
@@
 
-With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two experimental declarative 3D Web publishing frameworks are available.
-In the following we briefly introduce these open-source frameworks, which support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
-This discussion is followed by the presentation of the so-called ``polyfill'' approach for the further X3DOM and XML3D development activities.
+With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two open-source frameworks are available, which are briefly introduced in this section.
+Both frameworks support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
+This discussion is followed by the presentation of the so-called ``polyfill'' approach for further development activities.
 
@@
   \centering
-  \includegraphics[width=0.9\columnwidth]{images/polyfill.png}
+  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
   \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
@@
 
-% TODO: don't cite this link but present the whole idea here!!!
+% Don't cite this link but present the whole idea here!
 % http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
-Recently, the group proposed to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS.
-etc.).
-The proposal aims to explain existing and possible integration levels (Figure \ref{fig:polyInt}) and finally to identify an integration level the group is aiming for.
-% again, this is still bla bla
-% instead, insert text from slides - see slide 22 ff.
+We also propose to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS, etc.) to explain existing and possible integration levels (Figure \ref{fig:polyInt} shows that currently only level 2, DOM integration, is possible), and to finally identify a suitable integration level for declarative 3D.
+However, the latter imposes several problems in that it requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug.
+However, integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API.
+
 It quickly became clear that a higher level of integration requires additional APIs in user agents.
@@
 
-TODO % see slides, where these are discussed starting with slide 11/12
+In this section we briefly outline our proposed essential elements for an upcoming Dec3D standard. First of all, a start element allows the transition from the HTML flow layout to 3D transformations:
 
-Dec3D Scene Graph
-Modern and minimalistic scene graph or tree
-traversal and culling
-creates mesh/material/transform bundles for render layer
-picking
-Supports graph and tree structures for data reuse
-API layer, no DOM integration!
+\small
+\begin{verbatim}
+<div>
+    <dec3d style="border: 1px solid black;">
+    ...
+    </dec3d>
+</div>
+\end{verbatim}
+\normalsize
 
-Essential Dec3D DOM Elements
-Focus on minimalistic element set, not usability
-Represents SG concepts as DOM elements
-Adds DOM, Events and CSS to SG
+The DOM integration obviously extends the HTML hierarchy.
+But this comes along with certain issues like the correct treatment of CSS inheritance.
+Therefore, CSS integration is another point in that existing CSS properties should be used where applicable.
+Transformations for instance can be represented using CSS 3D Transforms \cite{webkit3DCSS}.
+In addition, a set of 3D-specific CSS properties (e.g.
+for defining the appearance) needs to be defined.
+However, custom CSS properties are not yet supported by any Web Browser, though finally this seems to change now \cite{CSSCustom}.
 
-Comprehensive Dec3D DOM Elements
-Builds on Essential and/or SG layer
-Focus on usability not minimalism
-Could build on existing SG standards (e.g. X3D) for convenience
+Moreover, not only standard DOM UI Events like `mouseover' and `click' shall be supported, but also events that contain 3D-specific context information.
+In this regard, the Browser API should expose convenience functions and special Dec3D base types consisting of a lean set of complex data types for 2D/3D graphics.
+Similarly, generic mesh data containers should be based on TypedArrays \cite{TypArr12} with appropriate interfaces to efficiently modify the data, where external binary containers are used as transmission format \cite{Behr12,Trevett12}.
+Thus, 3D geometry definition will stay close to OpenGL primitives.
 
-1. Start element
-Transition from HTML flow layout to 3D transform
+Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards like X3D \cite{Web3D-X3D}.
 
-2. CSS integration
-Use exiting CSS properties where applicable
-Define a set of 3D-specific CSS properties
-Needs to be supported in UAs!
+Essential DOM elements are a basic set of built-in shaders, which should be enough for most use cases.
+Advantages are that they do not entail timing attack issues and that they can adapt to any target device.
+As textures, HTML elements like \verb|<img>|, \verb|<video>|, \verb|<canvas>| shall be used, whereas elements like \verb|<svg>| and \verb|<html>| can serve as interactive textures.
 
-3. Hierarchy
-HTML hierarchy
-Comes with issues (e.g. CSS inheritance)
-
-4. Transformations
-Use CSS 3D Transforms
-
-5. Generic Data Containers
-Take modern HW into account
-Name: Type, TupleSize, Value
-Based on TypedArrays with interface to modify data
-
-6. External Data Containers
-External (binary) blobs / transmission format
-Maps to generic data containers
-
-7. Operators on Data Containers
-Allows dynamic meshes, image processing, compression
-Data driven approach: Xflow, X3D Routes, SVG Filter FX
-
-8. Shaders
-2-3 powerful fixed-function shaders
-Enough for many use-cases
-No timing attack issues
-Can adapt to target device
-
-9. Textures
-Use HTML elements: \verb|<img>|, \verb|<video>|, \verb|<canvas>|
-Interactive textures: \verb|<svg>|, \verb|<html>|
-
-10. 3D Geometry?
-Close to OpenGL primitives
-
-11. Lights
-Shader parameters in GL, but don't obstruct future concepts
-Intuitive way to define lights
-
-12. Viewpoints?
-Link to viewpoints (intra and inter-document)
-
-13. Camera navigation
-Pre-defined navigation modes?
-
-14. DOM Events
-Support standard events: mouseover, click
-3D-specific DOM Events contain 3D-specific context information
-
-15. Interfaces and Types?
-Provide IDLs for new elements
-Convenience functions
-Dec3D Base Types: Lean set of complex data types for 2D/3D
-Provide / use types that help to manage 3D scene, e.g. CSS Matrix
+Lights are basically also shader parameters in OpenGL/WebGL \cite{WebGL12}, but here shall be represented as own elements to provide intuitive ways to define lights and to not obstruct more advanced future concepts.
+Likewise, viewpoints can serve as link to certain points of interest within the document.
+However, it is still discussable if some pre-defined camera navigation modes should be part of the essential profile.
 
@@
 In this section we shape the agenda and identify upcoming research issues for the next development stage of Dec3D.
-During the 1\textsuperscript{st} International Workshop on Declarative 3D for the Web Architecture \cite{Dec3D2012} as well as during more informal group meetings at Web3D and SIGGRAPH in 2011/2012, the members and supporters of the Declarative 3D W3C Community Group agreed that what is needed to make the effort successful and for the W3C to adopt/develop a Dec3D standard include the following topics.
+During the 1\textsuperscript{st} Intl.
+Workshop on Declarative 3D \cite{Dec3D2012} as well as during more informal meetings at Web3D and SIGGRAPH in 2011/2012, the members and supporters of the Declarative 3D W3C Community Group agreed upon the following topics to make the effort successful and for the W3C to adopt/develop a Dec3D standard.
 

### 2013-03-11 Web3D2013-Dec3D/paper.tex
req update

@@
 Polyfill-based approaches allow deriving hard requirements for related and utilized W3C standards and user agent APIs immediately.
+
+\begin{description}
+\item[DOM] The Layer must be able to access and monitor changes in related DOM elements. This is currently achieved using the DOM Level 2 MotationEvents but may build on the new MutationObserver 
+\item[Events] The UA must support registration, firing and extending UI-Events. This is mainly done through dynamic property changes in the JavaScript object which represents the Event today
+\item[Custom CSS properties] The goal is to support scene management though custom CSS properties.
+This was not possible thought the last years but my change with the new CSS Custom Properties specification \cite{CSSCustom}.
+\item[CSS 3D Transform] CSS 3D Transforms module is now supported in all major browser.
+The methods to monitor final matrix changes are limited right now and should be extended for optimal performance
+\item[TypedArrays] TypedArrays where first introduces with the WebGL specification but are now a established method to process large portions of typed data efficiently.    
+\item[Generic GPU access] For client side rendering the Polyfill layer has to access the GPU functionally almost directly. WebGL is there a perfect candidate and Flash/Stag3D a second option. 
+\end{description}
+
 This leads to much more evolving concepts and solutions in contrast to an overall declarative 3D specification.

### 2013-03-11 Web3D2013-Dec3D/paper.tex
corrections

--- 2013-03-09 Web3D2013-Dec3D/paper.tex
@@
 
-Polyfill-based approaches allow deriving hard requirements for related and utilized W3C standards and user agent APIs immediately.
+Polyfill-based approaches allow deriving hard requirements for related and utilized W3C standards and user agent (UA) APIs immediately.
 This leads to much more evolving concepts and solutions in contrast to an overall declarative 3D specification.
+In the following we hence list our most important UA requirements.
+
+\begin{description}
+\item[DOM] The polyfill layer must be able to access and monitor changes in related DOM elements.
+This can be achieved using the now deprecated DOM Mutation Events as well as the new \emph{MutationObserver}\footnote{\url{https://dvcs.w3.org/hg/domcore/raw-file/tip/Overview.html\#mutation-observers}} objects.
+\item[Events] The UA must support registration, firing, and extending UI events.
+This is mainly done through dynamic property changes in the JavaScript object, which currently represent the event for performance reasons.
+\item[Custom CSS properties] The goal is to support scene management though custom CSS properties.
+Unfortunately, this was not possible in the last years, but may change with the new CSS Custom Properties specification \cite{CSSCustom}.
+\item[CSS 3D Transforms] The CSS 3D Transforms module \cite{webkit3DCSS} is now supported in all major browsers, but the methods to monitor final matrix changes are limited right now and should be further extended for optimal performance.
+\item[TypedArrays] TypedArrays \cite{TypArr12} were first introduced with the WebGL specification but are now an established method to process large portions of typed data efficiently.
+\item[Generic GPU access] For client-side rendering the polyfill layer has to access the GPU functionality almost directly.
+Here, WebGL is there a perfect candidate and Flash 11/Stage3D a second fallback option.
+\end{description}
+
 To sum up, the X3DOM and XML3D experimental declarative 3D Web publishing frameworks are designed to explore different options for adding 3D graphics to HTML.

### 2013-03-11 Web3D2013-Dec3D/paper.tex
added online id
Extended XML3D section.

@@
 
-\TOGonlineid{XXX} %TODO
+\TOGonlineid{paper1037} %TODO
 
@@
 
-DFKI's XML3D \cite{Sons2010} is a similar approach.
-In contrast to X3DOM, this format proposes a new tag set that is not based on X3D anymore.
-Furthermore, it does not define a scene-graph but a strict tree structure by means of a prototyping mechanism.
-In recent versions of XML3D, mesh data can also be externalized from the HTML document by storing it in separate JSON files.
-Moreover, the developers are working on a system for dynamic data-flow to efficiently support animations \cite{KleinSJRSB12}.
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{images/museum_with_tags2.pdf}
+  \caption{A highly dynamic declarative 3D scene with nine animated and skinned characters.
+The calculation is either performed with classical sequential JavaScript or in parallel exploiting Intel's River Trail proposal for parallel data computations in JavaScript.}
+  \label{fig:xflow}
+  \vspace{-0.2cm}
+\end{figure}
+
+XML3D \cite{Sons2010} is a collaboration of DFKI and Intel VCI.
+Similar to X3DOM, it employs CSS3, DOM scripting, DOM events, and WebGL.
+In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch.
+It does not define a scene-graph but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data.
+These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g.
+as JSON files.
+
+This generic and consistent handling of resources allows fine-granular composition of data sources.
+It can be combined with Xflow, a declarative data flow component~\cite{KleinSJRSB12} that allows for dynamic meshes, morphing, animation of shader parameters, image processing and AR.
+The data flow computation can be mapped to hardware, e.g.
+by composition of WebGL shaders or by exploiting available APIs such as Web Workers~\cite{w3c:webworkers} or River Trail~\cite{Herhut:2012}.
+Figure~\ref{fig:xflow} shows a XML3D scene with nine characters, skinned and animated with Xflow.
 

### 2013-03-11 Web3D2013-Dec3D/paper.tex
Added RTSG paper to platform independence section.

@@
 \paragraph{Platform Independence}
-3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only \cite{Schwenk12}.
+3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only \cite{Rubinstein:2009,Schwenk12}.
 This should allow for content to be portable across browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible.

### 2013-03-11 Web3D2013-Dec3D/paper.tex
little reformatting

@@
 \subsection{X3DOM and XML3D}
-
 With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two open-source frameworks are available, which are briefly introduced in this section.
@@
 Additionally, instead of implementing the somehow odd X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.
+
+\vspace{-1px}
 
@@
   \label{fig:xflow}
-  \vspace{-0.2cm}
+  \vspace{-0.3cm}
 \end{figure}
+
+\vspace{-1px}
 
@@
 These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g.
-as JSON files.
+as JSON.
+% files.
+
+\vspace{-1px}
 
@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=1.0\textwidth]{images/cad_apps.png}
-  \caption{From left to right: web-based design review application with annotation markers; interactive explosion of CAD model on iPad; lightweight web-based viewer for 3D CAD data; visualization and interactive exploration of simulation data by using extended mouse events.}
-  \label{fig:cadApps}
-  \vspace{-0.2cm}
-\end{figure*}
-
 \subsection{Polyfill Approach}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
-  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
-  \label{fig:polyInt}
-  \vspace{-0.2cm}
-\end{figure}
-
 % Don't cite this link but present the whole idea here!
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=1.0\textwidth]{images/cad_apps.png}
+  \caption{From left to right: web-based design review application with annotation markers; interactive explosion of CAD model on iPad; lightweight web-based viewer for 3D CAD data; visualization and interactive exploration of simulation data by using extended mouse events.}
+  \label{fig:cadApps}
+  \vspace{-0.2cm}
+\end{figure*}
+
 \begin{description}
@@
 \end{description}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
+  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
+  \label{fig:polyInt}
+  \vspace{-0.2cm}
+\end{figure}
 

### 2013-04-23 Web3D2013-Dec3D/paper.tex
switched to final mode

--- 2013-03-11 Web3D2013-Dec3D/paper.tex
@@
-\documentclass[review]{acmsiggraph}
+\documentclass{acmsiggraph}
+%\documentclass[review]{acmsiggraph}
 
@@
 
+\setlength{\ACMcopyrightspace}{1.0in}
+
 %\title{A Polyfill Approach to Declarative Integration of Interactive 3D Graphics\\ into the World-Wide Web}
-\title{Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda
-\\~} %some space that we don't get over 4(?) pages in the final version
+\title{Declarative Integration of Interactive 3D Graphics into the World-Wide Web:\\Principles, Current Approaches, and Research Agenda}
 
-\author{Name Lastname\thanks{e-mail:name.lastname@someemail.com}\\Some Research Institution}
-\pdfauthor{Name Lastname}
+\author{
+Jacek Jankowski\thanks{e-mail:jacek.jankowski@inria.fr} \\ INRIA \\ Bordeaux \\ France
+\and Sandy Ressler\thanks{e-mail:sandynist@gmail.com} \\ NIST \\ Gaithersburg MD \\ USA
+\and Kristian Sons\thanks{e-mail:kristian.sons@dfki.de} \\ DFKI GmbH \\ Saarbr\"ucken \\ Germany
+\and Yvonne Jung\thanks{e-mail:yvonne.jung@igd.fraunhofer.de} \\ Fraunhofer IGD \\ Darmstadt \\ Germany
+\and Johannes Behr\thanks{e-mail:johannes.behr@igd.fraunhofer.de} \\ Fraunhofer IGD \\ Darmstadt \\ Germany
+\and Philipp Slusallek\thanks{e-mail:slusallek@cs.uni-saarland.de} \\ DFKI GmbH \& Universit\"at des Saarlandes \\ Saarbr\"ucken \\ Germany
+}
+
+\pdfauthor{Jacek Jankowski, Sandy Ressler, Kristian Sons, Yvonne Jung, Johannes Behr, Philipp Slusallek}
 
@@
 \keywordlist
+
 \copyrightspace
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=1.0\columnwidth]{images/Dec3D-Architecture.png}
-  \caption{Proposed declarative 3D ``polyfill'' runtime architecture.}
-  \label{fig:polyArch}
-  \vspace{-0.2cm}
-\end{figure}
-
 \section{Declarative 3D Frameworks}
@@
 We also need to reach out to Web developers who could provide us with valuable feedback as early as possible.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=1.0\linewidth]{images/museum_with_tags2.pdf}
+  \caption{A highly dynamic declarative 3D scene with nine animated and skinned characters.
+The calculation is either performed with classical sequential JavaScript or in parallel exploiting Intel's River Trail proposal for parallel data computations in JavaScript.}
+  \label{fig:xflow}
+  \vspace{-0.2cm}
+\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=1.0\linewidth]{images/museum_with_tags2.pdf}
-  \caption{A highly dynamic declarative 3D scene with nine animated and skinned characters.
-The calculation is either performed with classical sequential JavaScript or in parallel exploiting Intel's River Trail proposal for parallel data computations in JavaScript.}
-  \label{fig:xflow}
-  \vspace{-0.2cm}
-\end{figure}
+It is also worth mentioning that the Web applications shown in Figures \ref{fig:chApps} and \ref{fig:cadApps} were not realized by graphics programmers but mainly by Web developers, using standard DOM scripting as well as JavaScript frameworks like jQuery \cite{jquery}.
+For example, to switch to a certain viewpoint when the user clicks onto the arrow button in the walkthrough application shown in Figure \ref{fig:chApps} (left), the application developer simply needs to update the 'bind' attribute of the "viewpoint" tag by calling its \emph{setAttribute} method.
 
@@
 These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g.
-as JSON files.
+as JSON.
+% files.
 
@@
 
-\begin{figure*}
-  \centering
-  \includegraphics[width=1.0\textwidth]{images/cad_apps.png}
-  \caption{From left to right: web-based design review application with annotation markers; interactive explosion of CAD model on iPad; lightweight web-based viewer for 3D CAD data; visualization and interactive exploration of simulation data by using extended mouse events.}
-  \label{fig:cadApps}
-  \vspace{-0.2cm}
-\end{figure*}
-
 \subsection{Polyfill Approach}
@@
   \centering
-  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
-  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
-  \label{fig:polyInt}
+  \includegraphics[width=1.0\columnwidth]{images/Dec3D-Architecture.png}
+  \caption{Proposed declarative 3D ``polyfill'' runtime architecture.}
+  \label{fig:polyArch}
   \vspace{-0.2cm}
@@
 Polyfills can also be used to add entirely new functionality to browsers.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
+  \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
+  \label{fig:polyInt}
+  \vspace{-0.2cm}
+\end{figure}
 
@@
 Here, Figure \ref{fig:polyArch} depicts our proposed declarative 3D polyfill runtime architecture, though we'd like to stress that the whole integration model is still evolving and open for discussion.
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=1.0\textwidth]{images/cad_apps.png}
+  \caption{From left to right: web-based design review application with annotation markers; interactive explosion of CAD model on iPad; lightweight web-based viewer for 3D CAD data; visualization and interactive exploration of simulation data by using extended mouse events.}
+  \label{fig:cadApps}
+  \vspace{-0.2cm}
+\end{figure*}
 

### 2013-04-25 Web3D2013-Dec3D/paper.tex
Switched position of Polyfill and Essential subsections and added some text to better link the subsections

@@
 As already mentioned, the Dec3D Community Group \cite{Dec3D} has been formed to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack, which will provide a foundation for future standardization.
-While this standardization is our goal, we still need platforms allowing for the experimentation and for the assessment of our design decisions.
+While this standardization is our goal, we still need platforms allowing for the experimentation and evaluation of our design decisions~(\ref{subsection:evaluation}).
 We also need to reach out to Web developers who could provide us with valuable feedback as early as possible.
+From the evaluations we extract the essential features~(\ref{subsection:essentials}) for a declarative approach.
+Finally, we check the level of integration we can achieve, emulating these features using existing technologies~(\ref{subsection:polyfill}).
 
@@
 
-\subsection{X3DOM and XML3D}
-
-With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two open-source frameworks are available, which are briefly introduced in this section.
-Both frameworks support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
-This discussion is followed by the presentation of the so-called ``polyfill'' approach for further development activities.
+\subsection{Evaluation platforms}
+\label{subsection:evaluation}
+With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two evaluation platforms are available to support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
 
@@
 
+Both platforms are available under an Open Source license and used in many national and international research projects.
+From the user feedback and use-cases we were able to collect from the evaluation platforms, we were able to derive the essential features, that both approaches have in common and that we propose for a Dec3D standard.
+
 % TODO: In the last paragraph of this section, we can write about the efforts to consolidate approaches and/or how the competing two frameworks influence one another for the common benefit...?
 This should be followed by the some link to the next section (polyfill).
+\subsection{Declarative 3D Essentials}
+\label{subsection:essentials}
+In this section we briefly outline our proposed essential elements for an upcoming Dec3D standard. First of all, a start element allows the transition from the HTML flow layout to 3D transformations:
+
+\small
+\begin{verbatim}
+<div>
+  <dec3d style="border: 1px solid black;">
+    ...
+  </dec3d>
+</div>
+\end{verbatim}
+\normalsize
+
+The DOM integration obviously extends the HTML hierarchy.
+But this comes along with certain issues like the correct treatment of CSS inheritance.
+Therefore, CSS integration is another point in that existing CSS properties should be used where applicable.
+Transformations for instance can be represented using CSS 3D Transforms \cite{webkit3DCSS}.
+In addition, a set of 3D-specific CSS properties (e.g.
+for defining the appearance) needs to be defined.
+However, custom CSS properties are not yet supported by any Web Browser, though finally this seems to change now \cite{CSSCustom}.
+
+Moreover, not only standard DOM UI Events like `mouseover' and `click' shall be supported, but also events that contain 3D-specific context information.
+In this regard, the Browser API should expose convenience functions and special Dec3D base types consisting of a lean set of complex data types for 2D/3D graphics.
+Similarly, generic mesh data containers should be based on TypedArrays \cite{TypArr12} with appropriate interfaces to efficiently modify the data, where external binary containers are used as transmission format \cite{Behr12,Trevett12}.
+Thus, 3D geometry definition will stay close to OpenGL primitives.
+
+Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards like X3D \cite{Web3D-X3D}.
+
+Essential DOM elements are a basic set of built-in shaders, which should be enough for most use cases.
+Advantages are that they do not entail timing attack issues and that they can adapt to any target device.
+As textures, HTML elements like \verb|<img>|, \verb|<video>|, \verb|<canvas>| shall be used, whereas elements like \verb|<svg>| and \verb|<html>| can serve as interactive textures.
+
+Lights are basically also shader parameters in OpenGL/WebGL \cite{WebGL12}, but here shall be represented as own elements to provide intuitive ways to define lights and to not obstruct more advanced future concepts.
+Likewise, viewpoints can serve as link to certain points of interest within the document.
+However, it is still discussable if some pre-defined camera navigation modes should be part of the essential profile.
 
 \subsection{Polyfill Approach}
+\label{subsection:polyfill}
 
@@
 
-\subsection{Declarative 3D Essentials}
-
-In this section we briefly outline our proposed essential elements for an upcoming Dec3D standard. First of all, a start element allows the transition from the HTML flow layout to 3D transformations:
-
-\small
-\begin{verbatim}
-<div>
-  <dec3d style="border: 1px solid black;">
-    ...
-  </dec3d>
-</div>
-\end{verbatim}
-\normalsize
-
-The DOM integration obviously extends the HTML hierarchy.
-But this comes along with certain issues like the correct treatment of CSS inheritance.
-Therefore, CSS integration is another point in that existing CSS properties should be used where applicable.
-Transformations for instance can be represented using CSS 3D Transforms \cite{webkit3DCSS}.
-In addition, a set of 3D-specific CSS properties (e.g.
-for defining the appearance) needs to be defined.
-However, custom CSS properties are not yet supported by any Web Browser, though finally this seems to change now \cite{CSSCustom}.
-
-Moreover, not only standard DOM UI Events like `mouseover' and `click' shall be supported, but also events that contain 3D-specific context information.
-In this regard, the Browser API should expose convenience functions and special Dec3D base types consisting of a lean set of complex data types for 2D/3D graphics.
-Similarly, generic mesh data containers should be based on TypedArrays \cite{TypArr12} with appropriate interfaces to efficiently modify the data, where external binary containers are used as transmission format \cite{Behr12,Trevett12}.
-Thus, 3D geometry definition will stay close to OpenGL primitives.
-
-Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards like X3D \cite{Web3D-X3D}.
-
-Essential DOM elements are a basic set of built-in shaders, which should be enough for most use cases.
-Advantages are that they do not entail timing attack issues and that they can adapt to any target device.
-As textures, HTML elements like \verb|<img>|, \verb|<video>|, \verb|<canvas>| shall be used, whereas elements like \verb|<svg>| and \verb|<html>| can serve as interactive textures.
-
-Lights are basically also shader parameters in OpenGL/WebGL \cite{WebGL12}, but here shall be represented as own elements to provide intuitive ways to define lights and to not obstruct more advanced future concepts.
-Likewise, viewpoints can serve as link to certain points of interest within the document.
-However, it is still discussable if some pre-defined camera navigation modes should be part of the essential profile.
-
 \section{Declarative 3D Agenda}

### 2013-04-25 Web3D2013-Dec3D/paper.tex
Minor update: INRIA to Inria
Apparently, Inria is not an acronym anymore... ;)
removed vspace

@@
 \author{
-Jacek Jankowski\thanks{e-mail:jacek.jankowski@inria.fr} \\ INRIA \\ Bordeaux \\ France
+Jacek Jankowski\thanks{e-mail:jacek.jankowski@inria.fr} \\ Inria \\ Bordeaux \\ France
 \and Sandy Ressler\thanks{e-mail:sandynist@gmail.com} \\ NIST \\ Gaithersburg MD \\ USA
@@
   \label{fig:DeclarativeVsImperative}
-  \vspace{-0.2cm}
+  %\vspace{-0.2cm}
 \end{figure}
@@
   \label{fig:chApps}
-  \vspace{-0.2cm}
+  %\vspace{-0.2cm}
 \end{figure*}
@@
 3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only \cite{Rubinstein:2009,Schwenk12}.
-This should allow for content to be portable across browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible.
+This should allow for content to be portable across web browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible.
 The results of rendering content under such different environments should be highly predictable.
@@
   \label{fig:xflow}
-  \vspace{-0.2cm}
+  %\vspace{-0.2cm}
 \end{figure}
@@
 \label{subsection:evaluation}
+
 With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two evaluation platforms are available to support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
@@
 
-Both platforms are available under an Open Source license and used in many national and international research projects.
-From the user feedback and use-cases we were able to collect from the evaluation platforms, we were able to derive the essential features, that both approaches have in common and that we propose for a Dec3D standard.
+Both platforms are available under an Open Source license and used in many national and international research and industrial projects.
+From the user feedback and use cases we have collected from both evaluation platforms, we were able to derive the essential features, that both approaches have in common and that we propose for a Dec3D standard.
 
-% TODO: In the last paragraph of this section, we can write about the efforts to consolidate approaches and/or how the competing two frameworks influence one another for the common benefit...?
-This should be followed by the some link to the next section (polyfill).
 \subsection{Declarative 3D Essentials}
 \label{subsection:essentials}
+
 In this section we briefly outline our proposed essential elements for an upcoming Dec3D standard. First of all, a start element allows the transition from the HTML flow layout to 3D transformations:
@@
   \label{fig:polyArch}
-  \vspace{-0.2cm}
+  %\vspace{-0.2cm}
 \end{figure}
@@
   \label{fig:polyInt}
-  \vspace{-0.2cm}
+  %\vspace{-0.2cm}
 \end{figure}
@@
   \label{fig:cadApps}
-  \vspace{-0.2cm}
+  %\vspace{-0.2cm}
 \end{figure*}

### 2013-04-26 Web3D2013-Dec3D/paper.tex
Changed 'DFKI GmbH' to just 'DFKI' and 'Universität des Saarlandes' to 'Saarland University'

@@
 \and Sandy Ressler\thanks{e-mail:sandynist@gmail.com} \\ NIST \\ Gaithersburg MD \\ USA
-\and Kristian Sons\thanks{e-mail:kristian.sons@dfki.de} \\ DFKI GmbH \\ Saarbr\"ucken \\ Germany
+\and Kristian Sons\thanks{e-mail:kristian.sons@dfki.de} \\ DFKI \\ Saarbr\"ucken \\ Germany
 \and Yvonne Jung\thanks{e-mail:yvonne.jung@igd.fraunhofer.de} \\ Fraunhofer IGD \\ Darmstadt \\ Germany
 \and Johannes Behr\thanks{e-mail:johannes.behr@igd.fraunhofer.de} \\ Fraunhofer IGD \\ Darmstadt \\ Germany
-\and Philipp Slusallek\thanks{e-mail:slusallek@cs.uni-saarland.de} \\ DFKI GmbH \& Universit\"at des Saarlandes \\ Saarbr\"ucken \\ Germany
+\and Philipp Slusallek\thanks{e-mail:slusallek@cs.uni-saarland.de} \\ DFKI \& Saarland University \\ Saarbr\"ucken \\ Germany
 }

### 2013-04-26 Web3D2013-Dec3D/paper.tex
First iteration over 'Declarative 3D Essentials'

@@
 
-In this section we briefly outline our proposed essential elements for an upcoming Dec3D standard. First of all, a start element allows the transition from the HTML flow layout to 3D transformations:
+The CG identified 15 key concepts of DOM-based Dec3D. In this section we briefly outline our proposed essential elements for an upcoming standard.
+
+A \textit{start element}~(1) marks the transition point from HTML box layout to 3D transformations.
+This element itself is integrated in the flow layout, fully styleable with CSS and behaves similar to e.g.
+the \verb|<canvas>| element.
+Within the 3D space we require a \textit{hierarchy}~(2) for 3D objects.
+HTML offers elements to structure documents hierarchically using elements such as \verb|<div>|, \verb|<span>|, \verb|<section>|, etc.
+These elements could be reused.
+On the other hand, HTML5 recently extended the set of structural elements to allow a more granular definition of document semantics.
+It could be useful to do the same for 3D scenes.
+Still open is the question, if the reuse of sub-graphs should be supported.
+This is a common feature in X3D and can also be found in SVG.
+On the other hand, having multiple paths to one element introduces issues, e.g.
+with CSS inheritance.
+
+CSS 3D Transforms~\cite{webkit3DCSS} should be used, to define \textit{transformations}~(3) on the hierarchical structure as well as on other 3D elements.
+This example illustrates the three later concepts:
 
@@
   <dec3d style="border: 1px solid black;">
+    <div style="transform: scale3d(2, 2, 2);">
       ...
+    </div>
   </dec3d>
@@
 
-The DOM integration obviously extends the HTML hierarchy.
-But this comes along with certain issues like the correct treatment of CSS inheritance.
-Therefore, CSS integration is another point in that existing CSS properties should be used where applicable.
-Transformations for instance can be represented using CSS 3D Transforms \cite{webkit3DCSS}.
+In general, a tight \textit{CSS integration}~(4) is a major concept: Existing CSS properties should be used where applicable (e.g.\ transform, opacity, color, etc.).
 In addition, a set of 3D-specific CSS properties (e.g.
@@
 
-Moreover, not only standard DOM UI Events like `mouseover' and `click' shall be supported, but also events that contain 3D-specific context information.
-In this regard, the Browser API should expose convenience functions and special Dec3D base types consisting of a lean set of complex data types for 2D/3D graphics.
-Similarly, generic mesh data containers should be based on TypedArrays \cite{TypArr12} with appropriate interfaces to efficiently modify the data, where external binary containers are used as transmission format \cite{Behr12,Trevett12}.
-Thus, 3D geometry definition will stay close to OpenGL primitives.
+For realtime interactive 3D graphics systems an \textit{event}~(5) model is essential.
+The Dec3D standard should leverage the DOM Event model where applicable.
+Not only standard DOM UI Events such as `mouseover' and `click' shall be supported, but also new events need to be defined to provide 3D-specific context information.
+In this regard, the Browser API should expose convenience functions and special Dec3D \textit{base types}~(6) consisting of a lean set of complex data types for 2D/3D graphics.
+In general, we aim a common set of base types usable not only for Dec3D, but also for WebGL API, Audio API, SVG, 3D Transforms etc.
 
-Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards like X3D \cite{Web3D-X3D}.
+Similarly, \textit{generic data containers}~(7) should be based on TypedArrays \cite{TypArr12} with appropriate interfaces for efficient modification of the data.
+These containers should be general enough to support meshes, animations, shaders, etc.
+This concept is very close to our efforts regarding \textit{external data containers}~(8) and their efficient transmission~\cite{Behr12,Trevett12}.
+Given that the generic data containers are close to buffer objects in recent graphics APIs, we aim \textit{3D geometry definitions}~(9) close to OpenGL primitives.
 
-Essential DOM elements are a basic set of built-in shaders, which should be enough for most use cases.
+Dec3D should support highly dynamic scenes.
+Using the DOM API to modify the scene data is not sufficient for many use-cases, e.g.\ dynamic meshes, image processing, compression, etc.
+Thus we need a concept to support efficient data processing that allows to map computations to available hardware.
+These should be expressed as \textit{operators}~(10) on the data containers above.
+
+Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards such as X3D ~\cite{Web3D-X3D} or COLLADA~\cite{Arnaud06}.
+
+Essential DOM elements are a basic set of pre-defined \textit{shaders}~(11), which should be enough for most use cases.
 Advantages are that they do not entail timing attack issues and that they can adapt to any target device.
-As textures, HTML elements like \verb|<img>|, \verb|<video>|, \verb|<canvas>| shall be used, whereas elements like \verb|<svg>| and \verb|<html>| can serve as interactive textures.
+However, the design should not obstruct programmable shaders in the future.
+HTML elements such as \verb|<img>|, \verb|<video>|, and \verb|<canvas>| can be used as \textit{textures}~(12), whereas elements such as \verb|<svg>| and \verb|<html>| can serve as interactive textures.
 
-Lights are basically also shader parameters in OpenGL/WebGL \cite{WebGL12}, but here shall be represented as own elements to provide intuitive ways to define lights and to not obstruct more advanced future concepts.
-Likewise, viewpoints can serve as link to certain points of interest within the document.
-However, it is still discussable if some pre-defined camera navigation modes should be part of the essential profile.
+\textit{Lights}~(13) are shader parameters in OpenGL/WebGL \cite{WebGL12}.
+In contrast, for Dec3D we aim a representation to provide a intuitive way to define lights and to not obstruct more advanced future concepts.
+Likewise, \textit{viewpoints}~(14) can serve as link to certain points of interest within the document.
+However, it is still discussable if some pre-defined \textit{camera navigation}~(15) modes should be part of the essential profile.
 

### 2013-04-26 Web3D2013-Dec3D/paper.tex
some words added...

@@
 This concept is very close to our efforts regarding \textit{external data containers}~(8) and their efficient transmission~\cite{Behr12,Trevett12}.
-Given that the generic data containers are close to buffer objects in recent graphics APIs, we aim \textit{3D geometry definitions}~(9) close to OpenGL primitives.
+Given that the generic data containers are close to vertex buffer objects in recent graphics APIs, we aim \textit{3D geometry definitions}~(9) close to OpenGL primitives.
 
@@
 However, the design should not obstruct programmable shaders in the future.
-HTML elements such as \verb|<img>|, \verb|<video>|, and \verb|<canvas>| can be used as \textit{textures}~(12), whereas elements such as \verb|<svg>| and \verb|<html>| can serve as interactive textures.
+HTML elements such as \verb|<img>|, \verb|<video>|, and \verb|<canvas>| can be used as \textit{textures}~(12), whereas elements such as \verb|<svg>| and \verb|<html>| could serve as interactive textures -- once the corresponding security issues are resolved.
 

### 2013-04-28 Web3D2013-Dec3D/paper.tex
some polishing

@@
 Hence, in this paper we explore the options for new declarative ways of incorporating 3D graphics directly into HTML to enable its use on any Web page.
-We present declarative 3D principles that guide the work of the \textit{Declarative 3D for the Web Architecture W3C Community Group} and describe the current state of the fundamentals to this initiative.
+We present declarative 3D principles that guide the work of the \emph{Declarative 3D for the Web Architecture W3C Community Group} and describe the current state of the fundamentals to this initiative.
 Finally, we draw an agenda for the next development stages of Declarative 3D for the Web.
@@
 The provided concepts must lift the hardware-oriented imperative application programming interfaces to an expressive and more easily usable level.
-Therefore not the low-level data structures of existing GPU layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights.
+Therefore, not the low-level data structures of existing GPU layers must be in the center of the design but high-level elements and items like 3D objects, transformations, material descriptions, and lights.
 Instead of teaching Web developers 3D graphics APIs, the goal is to bring 3D graphics to the point where it is natural for Web developers to just make use of it.
-While this might not be possible for every possible use of a low-level API, it still can cover the vast majority of use cases.
+While this might not be feasible for every possible use of a low-level API, it still can cover the vast majority of use cases.
 
@@
 The core mission is to determine the requirements, options, and use cases for the integration of interactive 3D graphics into the Web technology stack in a declarative way, which hopefully will provide a foundation for future standardization.
-Therefore, the group aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) \cite{EXI} and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest.
+
+The group thus aims at presenting common use cases that define how 3D might intersect and interact with HTML5, DOM events, CSS, SVG, GeoLocation, Augmented Reality, Efficient XML Interchange (EXI) \cite{EXI} and other key working groups, whereas certain complex data types (e.g., transformation matrices) and computations are also of mutual interest.
 In this regard, this paper presents the current state and efforts concerning Dec3D.
@@
 \paragraph{Platform Independence}
-3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only \cite{Rubinstein:2009,Schwenk12}.
+3D content needs to be described in a way that does rely neither on a specific render API such as OpenGL or DirectX nor on a specific rendering technique such as rasterization or ray tracing only (cp.
+e.g.
+\cite{Rubinstein:2009,Schwenk12}).
 This should allow for content to be portable across web browsers, rendering techniques, and hardware platforms, while taking advantage of available features wherever possible.
@@
 
+\subsection{Evaluation platforms}
+\label{subsection:evaluation}
+
+With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two evaluation platforms are available to support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
+
+Fraunhofer IGD's \emph{X3DOM} \cite{Behr2009,Behr12} is a JavaScript-based open-source framework for declarative 3D graphics in HTML5 that aims at extending the HTML DOM tree with declarative 3D objects while employing modern Web technologies like CSS3, Ajax, DOM scripting, as well as WebGL and -- as fallback -- Adobe's Flash 11 with Stage 3D \cite{flash} for GPU-accelerated rendering.
+The proposed 3D elements are mostly based upon the open ISO standard X3D \cite{Web3D-X3D}, though X3DOM introduces a special HTML profile that basically extends the X3D Interchange profile.
+Additionally, instead of implementing the somehow odd X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.
+
+Furthermore, to overcome various problems that come along when embedding 3D mesh data, which typically consists of several megabytes of vertex attribute data, directly into the DOM tree, the developers are working on efficient 3D mesh encodings that allow separating the node structure from the raw vertex data \cite{Behr12}.
+In this regard, Figures \ref{fig:chApps} and \ref{fig:cadApps} show several X3DOM applications from the cultural heritage and engineering domain, where the ability to efficiently handle big 3D data sets is of high importance.
+Based on this work, 3D transmission formats are now also a major topic at the Khronos Group \cite{Trevett12}, where it is similarly argued that raw 3D data, just like image, video or audio data, needs to be externalized from the HTML document.
+
+It is also worth mentioning that the Web applications shown in Figures \ref{fig:chApps} and \ref{fig:cadApps} were not realized by graphics programmers but mainly by Web developers, using standard DOM scripting as well as JavaScript frameworks like jQuery \cite{jquery}.
+For example, to switch to a certain viewpoint when the user clicks onto the arrow button in the walkthrough application shown in Figure \ref{fig:chApps} (left), the application developer simply needs to update the 'bind' attribute of the "viewpoint" tag by calling its \emph{setAttribute} method.
+Likewise, when selecting an element in the browser-based CAD and product structure viewer shown in the middle of Figure \ref{fig:cadApps}, the corresponding parts in the tree view are highlighted and vice versa.
+
+\emph{XML3D} \cite{Sons2010} is a collaboration of DFKI and Intel VCI.
+Similar to X3DOM, it employs CSS3, DOM scripting, DOM events, and WebGL.
+In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch.
+It does not define a scene-graph but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data.
+These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g.
+as JSON.
+
+This generic and consistent handling of resources allows fine-granular composition of data sources.
+It can be combined with Xflow, a declarative data flow component~\cite{KleinSJRSB12} that allows for dynamic meshes, morphing, animation of shader parameters, image processing and Augmented Reality.
+The data flow computation can be mapped to hardware, e.g.
+by composition of WebGL shaders or by exploiting available APIs such as Web Workers~\cite{w3c:webworkers} or River Trail~\cite{Herhut:2012}.
+Figure~\ref{fig:xflow} shows a XML3D scene with nine characters, skinned and animated with Xflow.
+
+Both platforms are available under an Open Source license and used in many national and international research and industrial projects.
+From the user feedback and use cases we have collected from both evaluation platforms, we were able to derive the essential features, that both approaches have in common and that we propose for a Dec3D standard.
+
 \begin{figure}
@@
 
-\subsection{Evaluation platforms}
-\label{subsection:evaluation}
-
-With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two evaluation platforms are available to support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
-
-Fraunhofer IGD's X3DOM \cite{Behr2009,Behr12} is a JavaScript-based open-source framework for declarative 3D graphics in HTML5 that aims at extending the HTML DOM tree with declarative 3D objects while employing modern Web technologies like CSS3, Ajax, DOM scripting, as well as WebGL and -- as fallback -- Adobe's Flash 11 with Stage 3D \cite{flash} for GPU-accelerated rendering.
-The proposed 3D elements are mostly based upon the open ISO standard X3D \cite{Web3D-X3D}, though X3DOM introduces a special HTML profile that basically extends the X3D Interchange profile.
-Additionally, instead of implementing the somehow odd X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.
-
-Furthermore, to overcome various problems that come along when embedding 3D mesh data, which typically consists of several megabytes of vertex attribute data, directly into the DOM tree, the developers are working on efficient 3D mesh encodings that allow separating the node structure from the raw vertex data \cite{Behr12}.
-In this regard, Figures \ref{fig:chApps} and \ref{fig:cadApps} show several X3DOM applications from the cultural heritage and engineering domain, where the ability to efficiently handle big 3D data sets is of high importance.
-Based on this work, 3D transmission formats are now also a major topic at the Khronos Group \cite{Trevett12}, where it is similarly argued that raw 3D data, just like image, video or audio data, needs to be externalized from the HTML document.
-
-It is also worth mentioning that the Web applications shown in Figures \ref{fig:chApps} and \ref{fig:cadApps} were not realized by graphics programmers but mainly by Web developers, using standard DOM scripting as well as JavaScript frameworks like jQuery \cite{jquery}.
-For example, to switch to a certain viewpoint when the user clicks onto the arrow button in the walkthrough application shown in Figure \ref{fig:chApps} (left), the application developer simply needs to update the 'bind' attribute of the "viewpoint" tag by calling its \emph{setAttribute} method.
-
-XML3D \cite{Sons2010} is a collaboration of DFKI and Intel VCI.
-Similar to X3DOM, it employs CSS3, DOM scripting, DOM events, and WebGL.
-In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch.
-It does not define a scene-graph but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data.
-These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g.
-as JSON.
-% files.
-
-This generic and consistent handling of resources allows fine-granular composition of data sources.
-It can be combined with Xflow, a declarative data flow component~\cite{KleinSJRSB12} that allows for dynamic meshes, morphing, animation of shader parameters, image processing and AR.
-The data flow computation can be mapped to hardware, e.g.
-by composition of WebGL shaders or by exploiting available APIs such as Web Workers~\cite{w3c:webworkers} or River Trail~\cite{Herhut:2012}.
-Figure~\ref{fig:xflow} shows a XML3D scene with nine characters, skinned and animated with Xflow.
-
-Both platforms are available under an Open Source license and used in many national and international research and industrial projects.
-From the user feedback and use cases we have collected from both evaluation platforms, we were able to derive the essential features, that both approaches have in common and that we propose for a Dec3D standard.
-
 \subsection{Declarative 3D Essentials}
@@
 
-The CG identified 15 key concepts of DOM-based Dec3D. In this section we briefly outline our proposed essential elements for an upcoming standard.
+The community group identified 15 key concepts of DOM-based Dec3D. In this section we briefly outline our proposed essential elements for an upcoming standard.
 
@@
 
-In general, a tight \textit{CSS integration}~(4) is a major concept: Existing CSS properties should be used where applicable (e.g.\ transform, opacity, color, etc.).
+In general, a tight \textit{CSS integration}~(4) is a major concept: existing CSS properties should be used where applicable (e.g., transform, opacity, color, etc.).
 In addition, a set of 3D-specific CSS properties (e.g.
@@
 
-For realtime interactive 3D graphics systems an \textit{event}~(5) model is essential.
+For real-time interactive 3D graphics systems an \textit{event}~(5) model is essential.
 The Dec3D standard should leverage the DOM Event model where applicable.
-Not only standard DOM UI Events such as `mouseover' and `click' shall be supported, but also new events need to be defined to provide 3D-specific context information.
+Not only standard DOM UI Events such as `mouseover' and `click' (with appropriate extensions for 3D scenes as already discussed in \cite{BehrJDA11}) shall be supported, but also new events need to be defined to provide 3D-specific context information.
 In this regard, the Browser API should expose convenience functions and special Dec3D \textit{base types}~(6) consisting of a lean set of complex data types for 2D/3D graphics.
@@
 Dec3D should support highly dynamic scenes.
-Using the DOM API to modify the scene data is not sufficient for many use-cases, e.g.\ dynamic meshes, image processing, compression, etc.
+Using the DOM API to modify the scene data is not sufficient for many use cases, e.g.\ dynamic meshes, image processing, compression, etc.
 Thus we need a concept to support efficient data processing that allows to map computations to available hardware.
 These should be expressed as \textit{operators}~(10) on the data containers above.
-
-Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards such as X3D ~\cite{Web3D-X3D} or COLLADA~\cite{Arnaud06}.
-
-Essential DOM elements are a basic set of pre-defined \textit{shaders}~(11), which should be enough for most use cases.
-Advantages are that they do not entail timing attack issues and that they can adapt to any target device.
-However, the design should not obstruct programmable shaders in the future.
-HTML elements such as \verb|<img>|, \verb|<video>|, and \verb|<canvas>| can be used as \textit{textures}~(12), whereas elements such as \verb|<svg>| and \verb|<html>| could serve as interactive textures -- once the corresponding security issues are resolved.
-
-\textit{Lights}~(13) are shader parameters in OpenGL/WebGL \cite{WebGL12}.
-In contrast, for Dec3D we aim a representation to provide a intuitive way to define lights and to not obstruct more advanced future concepts.
-Likewise, \textit{viewpoints}~(14) can serve as link to certain points of interest within the document.
-However, it is still discussable if some pre-defined \textit{camera navigation}~(15) modes should be part of the essential profile.
-
-\subsection{Polyfill Approach}
-\label{subsection:polyfill}
 
@@
 
-% Don't cite this link but present the whole idea here!
+Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards such as X3D ~\cite{Web3D-X3D} or COLLADA~\cite{Arnaud06}.
+
+Essential DOM elements include a basic set of pre-defined \textit{shaders}~(11), which should be enough for typical use cases.
+Advantages are that they do not entail timing attack issues and that they can adapt to any target device.
+However, the design should not obstruct programmable shaders in the future.
+HTML elements such as \verb|<img>|, \verb|<video>|, and \verb|<canvas>| can be used as \textit{textures}~(12), whereas elements such as \verb|<svg>| and \verb|<html>| could serve as interactive textures -- once the corresponding security issues are resolved.
+
+\textit{Lights}~(13) are shader parameters in OpenGL/ WebGL \cite{WebGL12}.
+In contrast, for Dec3D we aim a representation to provide a intuitive way to define lights (and thereby also shadows) and to not obstruct more advanced future concepts.
+Likewise, \textit{viewpoints}~(14) can serve as link to certain points of interest within the document.
+However, it is still discussable if some pre-defined \textit{camera navigation}~(15) modes -- to get directly started without having to modify the camera matrix with a special script -- should be part of the essential profile.
+
+\subsection{Polyfill Approach}
+\label{subsection:polyfill}
+
 % http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
-We also propose to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS, etc.) to explain existing and possible integration levels (Figure \ref{fig:polyInt} shows that currently only level 2, DOM integration, is possible), and to finally identify a suitable integration level for declarative 3D.
-However, the latter imposes several problems in that it requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug.
+We also propose to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS, etc.) to explain existing and possible integration levels (Figure \ref{fig:polyInt} shows that currently only level 2, DOM integration, is possible here), and to finally identify a suitable integration level for declarative 3D.
+The latter imposes several problems in that it requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug.
 However, integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API.
@@
 It quickly became clear that a higher level of integration requires additional APIs in user agents.
-After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a polyfill approach for the further development activities.
+After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a so-called polyfill approach for the further development activities.
 Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.
@@
   \centering
-  \includegraphics[width=0.85\columnwidth]{images/polyfill.png}
+  \includegraphics[width=0.9\columnwidth]{images/polyfill.png}
   \caption{Levels of integration: current browser APIs only allow integrating with DOM and DOM Events, but not yet fully with CSS.}
@@
 \paragraph{Clear Definition of Use Cases and Requirements}
-The group needs to agree on a collection of use cases, where embedding 3D data in HTML using a declarative approach provides significant benefit.
+The group needs to agree on a collection of relevant use cases, where embedding 3D data in HTML using a declarative approach provides significant benefit.
 Here, declarative approaches esp.
-seem relevant for the industry, since they tend to think in formats not in APIs.
+seem relevant for the industry, since they tend to think in formats (such as STL, STEP/IGES or CATIA) not in APIs.
 Each use case should explore how publishers and consumers benefit from Dec3D.
@@
 \paragraph{Outreach and Exemplar Applications}
-The group needs to continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the open-source frameworks X3DOM and XML3D.
-Therefore, several applications should be identified, each requiring and demonstrating different capabilities of Dec3D.
-For example, one application could require huge 3D datasets which are impractical for inclusion directly into the DOM; another could require real-time control of an external system (inter-process communication); another could integrate with a complex data base; another an illustration of complex data (information visualization) without a physical analog, and another a more traditional scientific visualization.
-Key to selection of these applications is the use and demonstration of a declarative 3D requirement.
+The group needs to further continue its outreach activities through the high quality demonstrations of the Declarative 3D philosophy using the open-source frameworks X3DOM and XML3D.
+Therefore, several applications have to be identified, each requiring and demonstrating different capabilities of Dec3D.
+For example, one application could require huge 3D datasets, which are impractical for inclusion directly into the DOM; another could require real-time control of an external system; another could integrate with a complex data base; another an illustration of complex data (information visualization) without a physical analog, and yet another a more traditional scientific visualization.
+Key to selection of these applications is the use and demonstration of a more declarative 3D requirement.
 
@@
 While WebGL, a 3D imperative graphics API in the Web context, is getting more and more traction, we are still missing an easy way to add interactive high-level declarative 3D objects to an HTML document to allow anyone to easily create, share, and experience interactive 3D graphics, with possibly wide ranging effects similar to those caused by the broad availability of video on the Web.
-The main motivation is thus to make it easy to add 3D to Web pages by bringing 3D to the Web developers and not vice versa.
+The main motivation is thus to make it easy to add 3D graphics to Web pages by bringing 3D to the Web developers and not vice versa.
 This can be achieved by fully integrating 3D content into HTML5 documents, where interactive 3D graphics is a first class DOM objects.

### 2013-04-29 Web3D2013-Dec3D/paper.tex
Polyfill section extended by Level of Integration

@@
 
-\subsection{Polyfill Approach}
+\subsection{Level of Integration and Polyfill Approach}
 \label{subsection:polyfill}
@@
 % http://www.w3.org/community/declarative3d/wiki/Level_of_Integration/
-We also propose to measure a level of integration for 3D graphics in terms of W3C technologies (DOM, CSS, etc.) to explain existing and possible integration levels (Figure \ref{fig:polyInt} shows that currently only level 2, DOM integration, is possible here), and to finally identify a suitable integration level for declarative 3D.
-The latter imposes several problems in that it requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug.
-However, integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API.
+We also propose a measure for the Level of Integration~(LOI) for 3D graphics into the W3C technology stack (DOM, CSS, etc.) to explain existing and possible integration levels~(Figure \ref{fig:polyInt}).
+Level 0 is the classical integration using plug-ins.
+Many X3D implementations as well as Adobe's Flash and Unity 3D use the plug-in model to integrate their rendering engine into the browser.
+LOI 1 provides a dedicated element in the DOM in combination with an API.
+Currently  WebGL and the \verb|<canvas>| provide this level of integration.
+The next level (LOI 2) has the scene description not in JavaScript but integrated in the DOM.
+Levels 3 and 4 provide an even better integration with CSS and easy debugging of 3D scenes.
+This requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug.
 
-It quickly became clear that a higher level of integration requires additional APIs in user agents.
-After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a so-called polyfill approach for the further development activities.
+Using existing APIs in the browser, we can emulate native Dec3D functionality up to level 2, with some initial support of existing CSS properties.
+Integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API, a higher level of integration requires additional APIs in user agents.
+After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a so-called \textit{polyfill} approach for the further development activities.
 Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.

### 2013-04-29 Web3D2013-Dec3D/paper.tex
Mention the two XML3D implementations, upper-cased all Polyfills and "Declarative 3D Polyfill Runtime Architecture" as a name.

@@
 \emph{XML3D} \cite{Sons2010} is a collaboration of DFKI and Intel VCI.
-Similar to X3DOM, it employs CSS3, DOM scripting, DOM events, and WebGL.
+Similar to X3DOM, it employs CSS3, DOM scripting, and DOM events.
 In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch.
@@
 Figure~\ref{fig:xflow} shows a XML3D scene with nine characters, skinned and animated with Xflow.
+
+There are two implementations of XML3D: The native implementation based on a  modified Chromium browser, and xml3d.js~\cite{sons:2013}, a Polyfill implementation based on WebGL and JavaScript.
 
@@
   \includegraphics[width=1.0\columnwidth]{images/Dec3D-Architecture.png}
-  \caption{Proposed declarative 3D ``polyfill'' runtime architecture.}
+  \caption{Proposed Declarative 3D ``Polyfill'' Runtime Architecture.}
   \label{fig:polyArch}
@@
 Integration levels 3 and 4 are currently not yet achievable by a JavaScript implementation, because the required APIs are missing: it is not possible to introduce new CSS properties or to extend developer tools by an API, a higher level of integration requires additional APIs in user agents.
-After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a so-called \textit{polyfill} approach for the further development activities.
+After discussions with browser vendors (namely Firefox and Chrome), who made it clear that integration of the desired extensions natively into their frameworks is not currently their priority, we decided to consider a so-called \textit{Polyfill} approach for the further development activities.
 Nevertheless, communicating and requesting additional and missing features like custom CSS properties or CSS monitoring is necessary.
 
-A \textit{polyfill} is a downloadable piece of code, which provides facilities that are not natively built-in to a Web browser \cite{Sharp2010}.
-For example, many features of HTML5 are not supported by versions of Internet Explorer older than version 8 or 9, but can be used by web pages if those pages install a so-called polyfill.
+A \textit{Polyfill} is a downloadable piece of code, which provides facilities that are not natively built-in to a Web browser \cite{Sharp2010}.
+For example, many features of HTML5 are not supported by versions of Internet Explorer older than version 8 or 9, but can be used by web pages if those pages install a so-called Polyfill.
 Polyfills can also be used to add entirely new functionality to browsers.
@@
 \begin{description}
-\item[DOM] The polyfill layer must be able to access and monitor changes in related DOM elements.
+\item[DOM] The Polyfill Layer must be able to access and monitor changes in related DOM elements.
 This can be achieved using the now deprecated DOM Mutation Events as well as the new \emph{MutationObserver}\footnote{\url{https://dvcs.w3.org/hg/domcore/raw-file/tip/Overview.html\#mutation-observers}} objects.
@@
 \item[TypedArrays] TypedArrays \cite{TypArr12} were first introduced with the WebGL specification but are now an established method to process large portions of typed data efficiently.
-\item[Generic GPU access] For client-side rendering the polyfill layer has to access the GPU functionality almost directly.
+\item[Generic GPU access] For client-side rendering the Polyfill Layer has to access the GPU functionality almost directly.
 Here, WebGL is there a perfect candidate and Flash 11/Stage3D a second fallback option.
@@
 To sum up, the X3DOM and XML3D experimental declarative 3D Web publishing frameworks are designed to explore different options for adding 3D graphics to HTML.
-Here, Figure \ref{fig:polyArch} depicts our proposed declarative 3D polyfill runtime architecture, though we'd like to stress that the whole integration model is still evolving and open for discussion.
+Here, Figure \ref{fig:polyArch} depicts our proposed Declarative 3D Polyfill Runtime Architecture, though we'd like to stress that the whole integration model is still evolving and open for discussion.
 

### 2013-04-29 Web3D2013-Dec3D/paper.tex
Uncommented Acknowledgements section and started with the DFKI Acknowledgements.
Update to Acknowledgements

@@
 
-%\section*{Acknowledgements}
+\section*{Acknowledgements}
+The research presented in this paper has been supported by the EU projects VERVE\footnote{http://www.verveconsortium.eu/} and FI-CONTENT, the BMBF project EMERGENT, the Intel Visual Computing Institute, the SFI's Lion2 and EI's Copernicus projects, and the Villes Transparentes project.
 

### 2013-04-29 Web3D2013-Dec3D/paper.tex
Minor updates

@@
 
-\emph{XML3D} \cite{Sons2010} is a collaboration of DFKI and Intel VCI.
-Similar to X3DOM, it employs CSS3, DOM scripting, and DOM events.
+\emph{XML3D} \cite{Sons2010} is a result of collaboration of DFKI and Intel VCI.
+Similarly to X3DOM, it employs CSS3, DOM scripting, and DOM events.
 In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch.
-It does not define a scene-graph but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data.
+It does not define a scene-graph, but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data.
 These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g.
@@
 
-There are two implementations of XML3D: The native implementation based on a  modified Chromium browser, and xml3d.js~\cite{sons:2013}, a Polyfill implementation based on WebGL and JavaScript.
-
+There are two implementations of XML3D: the native implementation based on a  modified Chromium browser, and xml3d.js~\cite{sons:2013}, a Polyfill implementation based on WebGL and JavaScript.
 Both platforms are available under an Open Source license and used in many national and international research and industrial projects.
@@
   \label{fig:xflow}
-  %\vspace{-0.2cm}
+  \vspace{-0.2cm}
 \end{figure}
@@
 
-Besides this, we further distinguish between essential Dec3D DOM Elements, were the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards such as X3D ~\cite{Web3D-X3D} or COLLADA~\cite{Arnaud06}.
+Besides this, we further distinguish between essential Dec3D DOM Elements, where the focus lies on a minimalistic element set, as well as more comprehensive Dec3D DOM Elements (see middle of Figure \ref{fig:polyArch}), which build on top of those with the focus on usability and which for convenience could build on existing scene-graph standards such as X3D ~\cite{Web3D-X3D} or COLLADA~\cite{Arnaud06}.
 

### 2013-04-29 Web3D2013-Dec3D/paper.tex
Further minor updates

@@
 
-As already mentioned, the Dec3D Community Group \cite{Dec3D} has been formed to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack, which will provide a foundation for future standardization.
+As already mentioned, the Declarative 3D Community Group \cite{Dec3D} has been formed to determine the requirements, options, and use cases for the declarative integration of interactive 3D graphics capabilities into the Web technology stack, which will provide a foundation for future standardization.
 While this standardization is our goal, we still need platforms allowing for the experimentation and evaluation of our design decisions~(\ref{subsection:evaluation}).
@@
 
-With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010} two evaluation platforms are available to support the ongoing discussion in the computer graphics and Web communities how an integration of HTML(5) and Declarative 3D content could look like.
+With X3DOM \cite{Behr2009} and XML3D \cite{Sons2010}, two evaluation platforms are available to support the ongoing discussion in the computer graphics and Web communities how an integration of HTML and declarative 3D content could look like.
 
@@
 The proposed 3D elements are mostly based upon the open ISO standard X3D \cite{Web3D-X3D}, though X3DOM introduces a special HTML profile that basically extends the X3D Interchange profile.
-Additionally, instead of implementing the somehow odd X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.
+Additionally, instead of implementing the X3D pointing device sensor component, X3DOM simply uses, and appropriately extends, the HTML UI/Mouse events such that 3D pick events are likewise supported.
 
-Furthermore, to overcome various problems that come along when embedding 3D mesh data, which typically consists of several megabytes of vertex attribute data, directly into the DOM tree, the developers are working on efficient 3D mesh encodings that allow separating the node structure from the raw vertex data \cite{Behr12}.
+Furthermore, to overcome various problems that come along when embedding 3D mesh data, which typically consists of several megabytes of vertex attribute data, directly into the DOM tree, efficient 3D mesh encodings are being developed that allow separating the node structure from the raw vertex data \cite{Behr12}.
 In this regard, Figures \ref{fig:chApps} and \ref{fig:cadApps} show several X3DOM applications from the cultural heritage and engineering domain, where the ability to efficiently handle big 3D data sets is of high importance.

### 2013-04-29 Web3D2013-Dec3D/paper.tex
added v-must

@@
 Jacek Jankowski\thanks{e-mail:jacek.jankowski@inria.fr} \\ Inria \\ Bordeaux \\ France
-\and Sandy Ressler\thanks{e-mail:sandynist@gmail.com} \\ NIST \\ Gaithersburg MD \\ USA
+\and Sandy Ressler\thanks{e-mail:sressler@acm.org} \\ NIST \\ Gaithersburg MD \\ USA
 \and Kristian Sons\thanks{e-mail:kristian.sons@dfki.de} \\ DFKI \\ Saarbr\"ucken \\ Germany
@@
   \includegraphics[width=1.0\columnwidth]{images/Dec3D-Architecture.png}
-  \caption{Proposed Declarative 3D ``Polyfill'' Runtime Architecture.}
+  \caption{The proposed Declarative 3D ``Polyfill'' Runtime Architecture.}
   \label{fig:polyArch}
@@
 \section*{Acknowledgements}
-The research presented in this paper has been supported by the EU projects VERVE\footnote{http://www.verveconsortium.eu/} and FI-CONTENT, the BMBF project EMERGENT, the Intel Visual Computing Institute, the SFI's Lion2 and EI's Copernicus projects, and the Villes Transparentes project.
+The research presented in this paper has been supported by the EU projects V-Must (\url{http://v-must.net}), VERVE (\url{http://www.verveconsortium.eu}) and FI-CONTENT, the BMBF project EMERGENT, the Intel Visual Computing Institute, the SFI's Lion2 and EI's Copernicus projects, and the Villes Transparentes project.
 

### 2013-04-29 Web3D2013-Dec3D/paper.tex
New image presenting debugging
Minor updates

@@
 
-It is also worth mentioning that the Web applications shown in Figures \ref{fig:chApps} and \ref{fig:cadApps} were not realized by graphics programmers but mainly by Web developers, using standard DOM scripting as well as JavaScript frameworks like jQuery \cite{jquery}.
+It is also worth mentioning that the Web applications shown in Figures \ref{fig:chApps} and \ref{fig:cadApps} were not realized by graphics programmers, but mainly by Web developers using standard DOM scripting as well as JavaScript frameworks like jQuery \cite{jquery}.
 For example, to switch to a certain viewpoint when the user clicks onto the arrow button in the walkthrough application shown in Figure \ref{fig:chApps} (left), the application developer simply needs to update the 'bind' attribute of the "viewpoint" tag by calling its \emph{setAttribute} method.
@@
 
-CSS 3D Transforms~\cite{webkit3DCSS} should be used, to define \textit{transformations}~(3) on the hierarchical structure as well as on other 3D elements.
-This example illustrates the three later concepts:
+CSS 3D Transforms~\cite{webkit3DCSS} should be used to define \textit{transformations}~(3) on the hierarchical structure as well as on other 3D elements.
+This example illustrates the three following concepts:
 
@@
 for defining the appearance) needs to be defined.
-However, custom CSS properties are not yet supported by any Web Browser, though finally this seems to change now \cite{CSSCustom}.
+However, custom CSS properties are not yet supported by any Web browser, though finally this seems to change now \cite{CSSCustom}.
 
@@
 Using the DOM API to modify the scene data is not sufficient for many use cases, e.g.\ dynamic meshes, image processing, compression, etc.
-Thus we need a concept to support efficient data processing that allows to map computations to available hardware.
+Thus, we need a concept to support efficient data processing that allows to map computations to available hardware.
 These should be expressed as \textit{operators}~(10) on the data containers above.
@@
 In contrast, for Dec3D we aim a representation to provide a intuitive way to define lights (and thereby also shadows) and to not obstruct more advanced future concepts.
-Likewise, \textit{viewpoints}~(14) can serve as link to certain points of interest within the document.
+Likewise, \textit{viewpoints}~(14) can serve as links to certain points of interest within the document.
 However, it is still discussable if some pre-defined \textit{camera navigation}~(15) modes -- to get directly started without having to modify the camera matrix with a special script -- should be part of the essential profile.
@@
 Levels 3 and 4 provide an even better integration with CSS and easy debugging of 3D scenes.
-This requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug.
+This requires having a set of 3D-specific CSS properties or even a full shader description via CSS as well as a deeper integration into debugging tools like Chrome Developer Tools or Firebug (for an example see Figure \ref{fig:debugger} and \cite{SonsSlusallekDec3D2012}).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\columnwidth]{images/museum.png}
+  \caption{LOI 3 in the native XML3D implementation: Debugging of a 3D scene using the Chrome developer toolbar. Note also that materials are attached to meshes using the 'shader' CSS property.}
+  \label{fig:debugger}
+  %\vspace{-0.2cm}
+\end{figure}
 
@@
 Participants should be willing to actively develop and donate materials towards the group's deliverables, as well as attend the group's teleconferences and face-to-face meetings.
+\vspace{-0.1cm}
 
@@
 From these use cases, the group needs to derive and prioritize the different required dimensions for the Dec3D technical specification \cite{JankowskiDec3D2012,LeFeuvreDec3D2012}.
+\vspace{-0.1cm}
 
@@
 Measurable properties need to be defined to quantitatively and/or qualitatively evaluate the achieved solution, document the pros and cons of each, and demonstrate that, based on the above analysis, there is a good chance of success in creating a W3C standard for Declarative 3D for the Web.
+\vspace{-0.1cm}
 
@@
 \section*{Acknowledgements}
-The research presented in this paper has been supported by the EU projects V-Must (\url{http://v-must.net}), VERVE (\url{http://www.verveconsortium.eu}) and FI-CONTENT, the BMBF project EMERGENT, the Intel Visual Computing Institute, the SFI's Lion2 and EI's Copernicus projects, and the Villes Transparentes project.
+The research presented in this paper has been supported by the EU projects V-Must (\url{http://v-must.net}), VERVE (\url{http://www.verveconsortium.eu}) and FI-CONTENT, the BMBF project EMERGENT, the Intel Visual Computing Institute, the SFI's Lion2 and EI's Copernicus (\url{http://copernicus.deri.ie}) projects, and the Villes Transparentes project.
 

### 2013-04-29 Web3D2013-Dec3D/paper.tex
final fixes

@@
 Similarly to X3DOM, it employs CSS3, DOM scripting, and DOM events.
-In contrast to X3DOM, XML3D is designed as a extension to HTML5 from scratch.
-It does not define a scene-graph, but uses the DOM tree structure for parent child relations and second level references e.g\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data.
+In contrast to X3DOM, XML3D is designed as an extension to HTML5 from scratch.
+It does not define a scene-graph, but uses the DOM tree structure for parent child relations and second level references e.g.\ via CSS to (re-)use resources such as geometry data, material descriptions or animation data.
 These resources are not necessarily in the same document, but can also come from external documents in various formats, e.g.
@@
 Figure~\ref{fig:xflow} shows a XML3D scene with nine characters, skinned and animated with Xflow.
+There are two implementations of XML3D: the native implementation based on a  modified Chromium browser, and xml3d.js~\cite{sons:2013}, a Polyfill implementation based on WebGL and JavaScript.
 
-There are two implementations of XML3D: the native implementation based on a  modified Chromium browser, and xml3d.js~\cite{sons:2013}, a Polyfill implementation based on WebGL and JavaScript.
-Both platforms are available under an Open Source license and used in many national and international research and industrial projects.
+The X3DOM and XML3D platforms are available under an Open Source license and used in many national and international research and industrial projects.
 From the user feedback and use cases we have collected from both evaluation platforms, we were able to derive the essential features, that both approaches have in common and that we propose for a Dec3D standard.
@@
   \centering
-  \includegraphics[width=0.9\columnwidth]{images/museum.png}
-  \caption{LOI 3 in the native XML3D implementation: Debugging of a 3D scene using the Chrome developer toolbar. Note also that materials are attached to meshes using the 'shader' CSS property.}
+  \includegraphics[width=0.95\columnwidth]{images/museum.png}
+  \caption{LOI 3 in the native XML3D implementation: debugging of a 3D scene using the Chrome developer toolbar. Note also that materials are attached to meshes using the 'shader' CSS property.}
   \label{fig:debugger}
@@
 To sum up, the X3DOM and XML3D experimental declarative 3D Web publishing frameworks are designed to explore different options for adding 3D graphics to HTML.
-Here, Figure \ref{fig:polyArch} depicts our proposed Declarative 3D Polyfill Runtime Architecture, though we'd like to stress that the whole integration model is still evolving and open for discussion.
+Here, Figure \ref{fig:polyArch} depicts our proposed Declarative 3D Polyfill Runtime Architecture, though we would like to stress that the whole integration model is still evolving and open for discussion.
 
